---
title: "Supplementary Figure 3: Pixel intensity profiles of Percoll density fractionation"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(tidyverse)
  library(cowplot)
})
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(f) file.path(params$data_dir, f)
out_path <- function(...) file.path(params$out_dir, ...)
```

## Figure legend

**Supplementary Figure 3. Density fractionation of proliferative and quiescent cultures in rich
and minimal media.** A) Images of density fractionation indicating the region assessed for pixel
quantification (yellow vertical line). B) Pixel intensity distribution across the analyzed region
with maximum distance at which peak pixel intensity was achieved is indicated (red line with red
text).

## Input data

| File | Description |
|---|---|
| `Rich_Exp_values.csv` | ImageJ line-profile (Distance (pixels), Gray_Value) along the marked region, rich media, proliferative |
| `Rich_Qui_values.csv` | Same, rich media, quiescent |
| `Minimal_Exp_values.csv` | Same, minimal media, proliferative |
| `Minimal_Qui_values.csv` | Same, minimal media, quiescent |

## Read profiles

```{r read}
files <- list(
  Rich_Exp    = "Rich_Exp_values.csv",
  Rich_Qui    = "Rich_Qui_values.csv",
  Minimal_Exp = "Minimal_Exp_values.csv",
  Minimal_Qui = "Minimal_Qui_values.csv"
)

data_list <- map(files, ~ read_csv(data_path(.x), show_col_types = FALSE))

# normalise column names to 'distance' and 'gray'
data_list <- map(data_list, function(df) {
  df <- df %>%
    rename_with(~ str_replace_all(.x, "[()]", "")) %>%
    rename_with(~ str_replace_all(.x, "[- ]", "_"))
  df %>%
    rename(
      distance = matches("(?i)distance"),
      gray     = matches("(?i)gray|(?i)grey|(?i)Gray_Value")
    )
})

# common axis limits
all_dist_range <- range(unlist(map(data_list, ~ range(.x$distance, na.rm = TRUE))))
all_gray_range <- range(unlist(map(data_list, ~ range(.x$gray, na.rm = TRUE))))
x_limits <- c(floor(all_dist_range[1]), ceiling(all_dist_range[2]))
y_limits <- c(floor(all_gray_range[1] - 1), ceiling(all_gray_range[2] + 1))

# distance of peak pixel intensity
imap_dfr(data_list, function(df, nm) {
  i <- which.max(df$gray)
  tibble(profile = nm, peak_distance = df$distance[i], peak_gray = df$gray[i])
}) %>% knitr::kable(digits = 2, caption = "Distance (pixels) at peak pixel intensity")
```

## Plot

```{r plot, fig.width=10, fig.height=8}
base_theme <- theme_classic(base_size = 12) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),
    plot.margin = margin(t = 18, r = 14, b = 12, l = 14)
  )

# line profile with a dashed vertical line and label at the peak
make_plot_mark_max <- function(df) {
  idx_max <- which.max(df$gray)
  x_max <- df$distance[idx_max]
  y_max <- df$gray[idx_max]
  x_text_offset <- (x_limits[2] - x_limits[1]) * 0.01

  ggplot(df, aes(x = distance, y = gray)) +
    geom_line(color = "black", linewidth = 0.7) +
    geom_vline(xintercept = x_max, linetype = "dashed", color = "red", linewidth = 0.6) +
    annotate("text", x = x_max + x_text_offset, y = y_max,
             label = paste0(round(x_max, 1)),
             hjust = 0, vjust = 10, size = 3.6, color = "red") +
    scale_x_continuous(limits = x_limits, expand = c(0, 0)) +
    scale_y_continuous(limits = y_limits, expand = c(0, 0)) +
    coord_cartesian(clip = "off") +
    base_theme
}

figure <- plot_grid(
  make_plot_mark_max(data_list$Rich_Exp),    make_plot_mark_max(data_list$Rich_Qui),
  make_plot_mark_max(data_list$Minimal_Exp), make_plot_mark_max(data_list$Minimal_Qui),
  ncol = 2, align = "hv", scale = 0.96
)

figure
ggsave(out_path("Supplementary_Figure_03_pixel_intensity.png"), figure,
       width = 10, height = 8, dpi = 300)
ggsave(out_path("Supplementary_Figure_03_pixel_intensity.pdf"), figure,
       width = 10, height = 8)
```

## Session info

```{r session}
sessionInfo()
```
