---
title: "Supplementary Figure 1: Cell cluster composition over time in rich and minimal media"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(tidyverse)
})
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(f) file.path(params$data_dir, f)
out_path <- function(...) file.path(params$out_dir, ...)
```

## Figure legend

**Supplementary Figure 1. Phase contrast microscopy images of rich and minimal media-grown
proliferative and quiescent cells used for bud index quantification.** A) Representative phase
contrast microscopy images showing cells at different growth phases grown in rich and minimal
media. Cells are counted and marked as 1 (singlet), 2 (doublet), 3 (triplet), 4 (quadruplet+) or
5 (ambiguous) and counted using Cell Counter plugin for ImageJ. B) Cell cluster type percentages
change with time in cells grown in rich and minimal media. Counts quantified from Cell Counter
outputs; minimum of 200 cells were counted per condition.

## Input data

| File | Description |
|---|---|
| `PC_Microscopy_analysis.csv` | ImageJ Cell Counter counts per phase-contrast image (Singlet, Doublet, Triplet, Quadruplet+, Ambiguous) for rich and minimal media at 2, 4, 6, 24, 48, 72 h |

## Aggregate counts

Counts are summed per Media × Time; ambiguous cells are excluded, so the four cluster types add
up to 100%.

```{r aggregate}
df <- read_csv(data_path("PC_Microscopy_analysis.csv"), show_col_types = FALSE)

quad_col <- grep("^Quadruplet", names(df), value = TRUE)[1]
if (!is.na(quad_col) && quad_col != "Quadruplet") {
  names(df)[names(df) == quad_col] <- "Quadruplet"
}

df <- df %>%
  mutate(across(c(Singlet, Doublet, Triplet, Quadruplet, Time), as.numeric),
         Media = as.factor(Media))

agg <- df %>%
  group_by(Media, Time) %>%
  summarise(
    Singlet = sum(Singlet, na.rm = TRUE),
    Doublet = sum(Doublet, na.rm = TRUE),
    Triplet = sum(Triplet, na.rm = TRUE),
    `Quadruplet+` = sum(Quadruplet, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(total_nonamb = Singlet + Doublet + Triplet + `Quadruplet+`) %>%
  filter(total_nonamb > 0) %>%
  arrange(Media, as.numeric(Time))

knitr::kable(agg, caption = "Summed counts per Media x Time (ambiguous excluded)")

agg_long <- agg %>%
  pivot_longer(cols = c(Singlet, Doublet, Triplet, `Quadruplet+`),
               names_to = "cluster", values_to = "count") %>%
  mutate(percent = round(count / total_nonamb * 100, 1)) %>%
  mutate(
    Time = factor(Time, levels = sort(unique(as.numeric(as.character(Time))))),
    cluster = factor(cluster, levels = c("Singlet", "Doublet", "Triplet", "Quadruplet+"))
  )

agg_long %>%
  select(Media, Time, cluster, percent) %>%
  pivot_wider(names_from = cluster, values_from = percent) %>%
  knitr::kable(caption = "Cluster type percentages (%)")
```

## Plot

```{r plot, fig.width=9, fig.height=4}
p <- ggplot(agg_long, aes(x = Time, y = percent, fill = cluster)) +
  geom_col(color = "black") +
  facet_wrap(~ Media, nrow = 1) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 100.5),
                     labels = function(x) paste0(x, "%")) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Time (h)", y = "Percentage of cells (%)", fill = "Cluster type",
       title = "Cell cluster composition by Media and Time") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(hjust = 0.5))

p
ggsave(out_path("Supplementary_Figure_01_cell_cluster_composition.png"), p,
       width = 9, height = 4, dpi = 300)
ggsave(out_path("Supplementary_Figure_01_cell_cluster_composition.pdf"), p,
       width = 9, height = 4)
```

## Session info

```{r session}
sessionInfo()
```
