---
title: "Figure 6: Viability of selected TF KOs at Day 3 and Day 7 of quiescence"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
  version_name: "OzanImir_20260401_v1"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(tidyverse)
  library(cowplot)
})
set.seed(1)
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(...) file.path(params$data_dir, "flow_validation", ...)
out_path <- function(...) file.path(params$out_dir, ...)
version_name <- params$version_name
```

## Figure legend

**Figure 6. Transcriptional regulation of quiescence.** A) Screen of TF KOs for decreased survival
in quiescence (validation of selected KOs; panel B in the current figure draft).

Nineteen TF KOs selected from the screen and the wild-type control (WT, SC5314) were grown to
quiescence in triplicate and measured by SYTO9/PI flow cytometry at Day 3 and Day 7.
Viability = Live cells / (Live cells + Dead cells). Bars show the mean of triplicates, error bars
the standard error, points the individual replicates. Each KO is compared with WT on the same day
by Welch's two-sample t-test with Benjamini-Hochberg correction within each day;
\* adjusted p < 0.05.

## Input data

All inputs are in `data/flow_validation/`:

| File | Description |
|---|---|
| `sample_sheet_vignette_simple.csv` | FCS file name, well, day, strain (orf19 ID) and knockout (gene name) |
| `Vignette-Experiment-Details.csv` | CytoExploreR experiment details (same content as the sample sheet) |
| `Vignette-Experiment-Markers.csv` | CytoExploreR channel/marker table (SYTO9 B3-A, PI YG4-A) |
| `cytek_gating_OzanImir_20260401_v1.csv` | CytoExploreR gating template (Cells, Single_cells, Live_cells, Dead_cells) |
| `viability_by_replicate_OzanImir_20260401_v1.csv` | Gated Live/Dead counts and viability per sample (120 samples) |
| `viability_summary_OzanImir_20260401_v1.csv` | Original mean/SE per day and KO (used for verification only) |
| `viability_vs_WT_significance_OzanImir_20260401_v1.csv` | Original t-test results (used for verification only) |

The raw FCS files are not distributed.

## Gating (not run: requires FCS files)

Gating was done with CytoExploreR on the FCS files (not distributed), using the gating template in
`data/flow_validation/`. Transforms: arcsinh, cofactor 150 for SYTO9 (B3-A in this experiment) and
PI (YG4-A), cofactor 500 for FSC/SSC. Gates: Cells (FSC-A/SSC-A) > Single_cells (SSC-A/SSC-H) >
Live_cells and Dead_cells (YG4-A/B3-A). The per-sample counts were saved as
`viability_by_replicate_OzanImir_20260401_v1.csv`.

```{r gating, eval = FALSE}
library(CytoExploreR)
library(flowWorkspace)

# FCS files are expected in data/flow_validation/FCS_files (not distributed)
my_gating_set <- cyto_setup(path = data_path("FCS_files"), restrict = TRUE, select = "fcs")
experiment_details <- read_csv(data_path("Vignette-Experiment-Details.csv"), show_col_types = FALSE)
for (i in seq_along(experiment_details)) {
  pData(my_gating_set)[names(experiment_details)[i]] <- experiment_details[[i]]
}

SYTO9_trans <- cyto_transformer_arcsinh(my_gating_set, channels = c("B3-A"), cofactor = 150)
PI_trans <- cyto_transformer_arcsinh(my_gating_set, channels = c("YG4-A"), cofactor = 150)
FSC_SSC_trans <- cyto_transformer_arcsinh(my_gating_set,
                                          channels = c("FSC-A", "FSC-H", "SSC-A", "SSC-H"),
                                          cofactor = 500)
combined_trans <- cyto_transformer_combine(SYTO9_trans, PI_trans, FSC_SSC_trans)
transformed_gating_set <- cyto_transform(my_gating_set, trans = combined_trans)

cyto_gatingTemplate_apply(transformed_gating_set,
                          gatingTemplate = data_path(paste0("cytek_gating_", version_name, ".csv")))

sample_sheet <- read_csv(data_path("sample_sheet_vignette_simple.csv"), show_col_types = FALSE)
sample_info <- pData(transformed_gating_set) %>%
  as.data.frame() %>%
  as_tibble() %>%
  mutate(sample_name = sampleNames(transformed_gating_set))
if (!"name" %in% names(sample_info)) {
  sample_info <- sample_info %>% mutate(name = basename(sample_name))
}
sample_info <- sample_info %>%
  select(sample_name, name) %>%
  left_join(sample_sheet, by = "name")

pop_counts <- gs_pop_get_stats(transformed_gating_set, path = "auto") %>%
  as_tibble() %>%
  rename(sample_name = sample, population = pop, cell_count = count) %>%
  mutate(population = sub(".*/", "", population)) %>%
  filter(population %in% c("Live_cells", "Dead_cells")) %>%
  select(sample_name, population, cell_count) %>%
  pivot_wider(names_from = population, values_from = cell_count,
              values_fill = list(cell_count = 0))

pop_counts %>%
  left_join(sample_info, by = "sample_name") %>%
  mutate(viability = Live_cells / (Live_cells + Dead_cells)) %>%
  write_csv(data_path(paste0("viability_by_replicate_", version_name, ".csv")))
```

## Viability per replicate

```{r viability}
sample_sheet <- read_csv(data_path("sample_sheet_vignette_simple.csv"), show_col_types = FALSE)

# Knockout order follows the plate layout
knockout_order <- sample_sheet %>%
  arrange(factor(day, levels = c("Day3", "Day7")), readr::parse_number(well)) %>%
  distinct(knockout) %>%
  pull(knockout)

viability_df <- read_csv(data_path(paste0("viability_by_replicate_", version_name, ".csv")),
                         show_col_types = FALSE) %>%
  mutate(day = factor(day, levels = c("Day3", "Day7")),
         knockout = factor(knockout, levels = knockout_order),
         viability = Live_cells / (Live_cells + Dead_cells))
stopifnot(nrow(viability_df) == 120L)

viability_summary <- viability_df %>%
  group_by(day, knockout) %>%
  summarise(n = dplyr::n(),
            mean_viability = mean(viability, na.rm = TRUE),
            se_viability = ifelse(dplyr::n() > 1,
                                  sd(viability, na.rm = TRUE) / sqrt(dplyr::n()),
                                  NA_real_),
            .groups = "drop")
write_csv(viability_summary, out_path("viability_summary.csv"))
knitr::kable(viability_summary, digits = 4)
```

## Statistics: each KO vs WT within each day

```{r significance}
wt_label <- "WT"

significance_df <- viability_df %>%
  filter(!is.na(day), !is.na(knockout), !is.na(viability)) %>%
  group_by(day, knockout) %>%
  summarise(values = list(viability), .groups = "drop") %>%
  left_join(viability_df %>%
              filter(knockout == wt_label, !is.na(viability)) %>%
              group_by(day) %>%
              summarise(wt_values = list(viability), .groups = "drop"),
            by = "day") %>%
  rowwise() %>%
  mutate(p_value = if (knockout == wt_label) {
    NA_real_
  } else {
    tryCatch(suppressWarnings(t.test(unlist(values), unlist(wt_values))$p.value),
             error = function(e) NA_real_)
  }) %>%
  ungroup() %>%
  group_by(day) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"),
         signif_label = case_when(knockout == wt_label ~ "",
                                  !is.na(p_adj) & p_adj < 0.05 ~ "*",
                                  TRUE ~ "")) %>%
  ungroup() %>%
  select(day, knockout, p_value, p_adj, signif_label)
write_csv(significance_df, out_path("viability_vs_WT_significance.csv"))
knitr::kable(significance_df, digits = 4)
```

Comparison with the saved results of the original analysis:

```{r verify}
orig_summary <- read_csv(data_path(paste0("viability_summary_", version_name, ".csv")),
                         show_col_types = FALSE)
orig_sig <- read_csv(data_path(paste0("viability_vs_WT_significance_", version_name, ".csv")),
                     show_col_types = FALSE, col_types = cols(signif_label = col_character()))

cmp_summary <- viability_summary %>%
  mutate(day = as.character(day), knockout = as.character(knockout)) %>%
  inner_join(orig_summary, by = c("day", "knockout"), suffix = c("", "_orig"))
cmp_sig <- significance_df %>%
  mutate(day = as.character(day), knockout = as.character(knockout)) %>%
  inner_join(orig_sig, by = c("day", "knockout"), suffix = c("", "_orig"))

data.frame(
  check = c("summary rows matched", "max |mean diff|", "max |SE diff|",
            "significance rows matched", "max |p diff|", "max |p_adj diff|", "stars identical"),
  value = c(nrow(cmp_summary),
            max(abs(cmp_summary$mean_viability - cmp_summary$mean_viability_orig)),
            max(abs(cmp_summary$se_viability - cmp_summary$se_viability_orig)),
            nrow(cmp_sig),
            max(abs(cmp_sig$p_value - cmp_sig$p_value_orig), na.rm = TRUE),
            max(abs(cmp_sig$p_adj - cmp_sig$p_adj_orig), na.rm = TRUE),
            all(coalesce(cmp_sig$signif_label, "") == coalesce(cmp_sig$signif_label_orig, "")))
)
```

## Figure panel: Day 3 and Day 7 viability of selected TF KOs

```{r figure-validation, fig.width = 16, fig.height = 6}
plot_df <- viability_summary %>%
  left_join(significance_df, by = c("day", "knockout")) %>%
  mutate(star_y = pmin(mean_viability + dplyr::coalesce(se_viability, 0) + 0.04, 1.05))

p_validation <- ggplot(plot_df, aes(x = knockout, y = mean_viability)) +
  geom_col(width = 0.75, fill = "#4C78A8", color = "black") +
  geom_errorbar(aes(ymin = pmax(mean_viability - se_viability, 0),
                    ymax = pmin(mean_viability + se_viability, 1)),
                width = 0.2, linewidth = 0.4) +
  geom_point(data = viability_df, aes(x = knockout, y = viability), inherit.aes = FALSE,
             alpha = 0.3, size = 2, position = position_jitter(width = 0.12, height = 0)) +
  geom_text(data = plot_df %>% filter(signif_label != ""),
            aes(x = knockout, y = star_y, label = signif_label), inherit.aes = FALSE,
            size = 5, fontface = "bold", vjust = 0) +
  facet_wrap(~day, nrow = 1) +
  scale_y_continuous(limits = c(0, 1.08), labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Knockout", y = "Viability", title = "KO Strain Collection Viability",
       subtitle = "Viability = Live_cells / (Live_cells + Dead_cells)",
       caption = "* adjusted p < 0.05 vs WT within the same day") +
  cowplot::theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        strip.background = element_rect(fill = "grey90", color = NA),
        strip.text = element_text(face = "bold"))
p_validation

ggsave(out_path("Figure_6_TFKO_viability_validation.pdf"), p_validation, width = 16, height = 6)
ggsave(out_path("Figure_6_TFKO_viability_validation.png"), p_validation, width = 16, height = 6,
       dpi = 300)
```

## Session info

```{r session-info}
sessionInfo()
```
