---
title: "Figure 6: Micafungin tolerance of quiescent TF KOs"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
  version_name: "OzanImir_20260809_v1"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(tidyverse)
  library(cowplot)
})
set.seed(1)
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(...) file.path(params$data_dir, "flow_micafungin_quiescent", ...)
out_path <- function(...) file.path(params$out_dir, ...)
version_name <- params$version_name
```

## Figure legend

**Figure 6. Transcriptional regulation of quiescence.** Viability of quiescent TF KOs after
micafungin treatment. Quiescent counterpart of panel C; not in the current figure draft (panel C
is the proliferative, normalized-to-untreated version, `Figure_6_TFKO_micafungin_proliferative.Rmd`).

Wild type (WT, SC5314), the parental background strain of the KO collection (BG) and the
knockouts of NRG1 (NRG, orf19.7150), NDT80 (NDT, orf19.2119) and STP4 (STP, orf19.909) were grown to
quiescence (Day 3) in rich media, treated with micafungin (Untreated, 0.001, 0.01, 0.1, 1 and
10 µg/mL) in triplicate and measured by SYTO9/PI flow cytometry.
Viability = Live cells / (Live cells + Dead cells). Bars show the mean of triplicates, error bars
the standard error, points the individual replicates. At each dose each KO strain is compared
with WT by a two-sided paired t-test (paired by replicate population, `repeat`), Holm-corrected
within each dose (4 tests); Untreated is not tested.
\* p < 0.05, \*\* p < 0.01, \*\*\* p < 0.001, \*\*\*\* p < 0.0001.

## Input data

All inputs are in `data/flow_micafungin_quiescent/`:

| File | Description |
|---|---|
| `sample_sheet_vignette_simple.csv` | FCS file name, knockout, media, micafungin dose and replicate (`repeat`) |
| `Vignette-Experiment-Details.csv` | CytoExploreR experiment details (sample sheet joined to file names) |
| `Vignette-Experiment-Markers.csv` | CytoExploreR channel/marker table (SYTO9 B7-A, PI YG4-A) |
| `cytek_gating_OzanImir_20260809_v1.csv` | CytoExploreR gating template (Cells, Single_cells, Live_cells, Dead_cells) |
| `viability_raw_by_strain_OzanImir_20260809_v1.csv` | Gated Live/Dead counts and viability per sample, rich media (90 samples) |
| `strain_vs_WT_paired_ttest_OzanImir_20260809_v1.csv` | Original paired t-test results (used for verification only) |

The raw FCS files are not distributed.

## Gating (not run: requires FCS files)

Gating was done with CytoExploreR on the FCS files (not distributed), using the gating template in
`data/flow_micafungin_quiescent/`. Transforms: arcsinh, cofactor 150 for SYTO9 (B7-A) and PI
(YG4-A), cofactor 500 for FSC/SSC. Gates: Cells (FSC-A/SSC-A) > Single_cells (SSC-A/SSC-H) >
Live_cells and Dead_cells (YG4-A/B7-A). The rich-media per-sample counts were saved as
`viability_raw_by_strain_OzanImir_20260809_v1.csv`.

```{r gating, eval = FALSE}
library(CytoExploreR)
library(flowWorkspace)

# FCS files are expected in data/flow_micafungin_quiescent/FCS_files (not distributed)
my_gating_set <- cyto_setup(path = data_path("FCS_files"), restrict = TRUE, select = "fcs")
experiment_details <- read_csv(data_path("Vignette-Experiment-Details.csv"), show_col_types = FALSE)
for (i in seq_along(experiment_details)) {
  pData(my_gating_set)[names(experiment_details)[i]] <- experiment_details[[i]]
}

SYTO9_trans <- cyto_transformer_arcsinh(my_gating_set, channels = c("B7-A"), cofactor = 150)
PI_trans <- cyto_transformer_arcsinh(my_gating_set, channels = c("YG4-A"), cofactor = 150)
FSC_SSC_trans <- cyto_transformer_arcsinh(my_gating_set,
                                          channels = c("FSC-A", "FSC-H", "SSC-A", "SSC-H"),
                                          cofactor = 500)
combined_trans <- cyto_transformer_combine(SYTO9_trans, PI_trans, FSC_SSC_trans)
transformed_gating_set <- cyto_transform(my_gating_set, trans = combined_trans)

cyto_gatingTemplate_apply(transformed_gating_set,
                          gatingTemplate = data_path(paste0("cytek_gating_", version_name, ".csv")))

sample_sheet <- read_csv(data_path("sample_sheet_vignette_simple.csv"), show_col_types = FALSE)
sample_info <- pData(transformed_gating_set) %>%
  as.data.frame() %>%
  as_tibble() %>%
  mutate(sample_name = sampleNames(transformed_gating_set))
if (!"name" %in% names(sample_info)) {
  sample_info <- sample_info %>% mutate(name = basename(sample_name))
}
sample_info <- sample_info %>%
  select(sample_name, name) %>%
  left_join(sample_sheet, by = "name")

pop_counts <- gs_pop_get_stats(transformed_gating_set, path = "auto") %>%
  as_tibble() %>%
  rename(sample_name = sample, population = pop, cell_count = count) %>%
  mutate(population = sub(".*/", "", population)) %>%
  filter(population %in% c("Live_cells", "Dead_cells")) %>%
  select(sample_name, population, cell_count) %>%
  pivot_wider(names_from = population, values_from = cell_count,
              values_fill = list(cell_count = 0))

pop_counts %>%
  left_join(sample_info, by = "sample_name") %>%
  mutate(strain = dplyr::recode(knockout, "WT" = "WT", "BG" = "BG", "NRG1" = "NRG",
                                "NDT80" = "NDT", "STP4" = "STP"),
         viability = Live_cells / (Live_cells + Dead_cells)) %>%
  filter(media == "Rich", !is.na(strain), !is.na(dose), !is.na(viability)) %>%
  write_csv(data_path(paste0("viability_raw_by_strain_", version_name, ".csv")))
```

## Viability per replicate

```{r viability}
strain_levels <- c("WT", "BG", "NRG", "NDT", "STP")
dose_levels <- c("Untreated", "0.001", "0.01", "0.1", "1", "10")

raw_df <- read_csv(data_path(paste0("viability_raw_by_strain_", version_name, ".csv")),
                   col_types = cols(dose = col_character(), `repeat` = col_integer()),
                   show_col_types = FALSE) %>%
  mutate(strain = factor(strain, levels = strain_levels),
         dose = factor(dose, levels = dose_levels),
         media = factor(media, levels = c("Rich", "Minimal")),
         viability = Live_cells / (Live_cells + Dead_cells)) %>%
  filter(media == "Rich", !is.na(strain), !is.na(dose), !is.na(viability))
stopifnot(nrow(raw_df) == 90L)

raw_summary <- raw_df %>%
  group_by(strain, dose) %>%
  summarise(n = dplyr::n(),
            mean_viability = mean(viability, na.rm = TRUE),
            se_viability = ifelse(dplyr::n() > 1,
                                  sd(viability, na.rm = TRUE) / sqrt(dplyr::n()),
                                  NA_real_),
            .groups = "drop")
write_csv(raw_summary, out_path("viability_raw_summary.csv"))
knitr::kable(raw_summary, digits = 4)
```

## Statistics: each KO strain vs WT at each dose

Replicates 1-3 come from the same three populations across all conditions, so tests are paired
by `repeat`. With n = 3 pairs a Wilcoxon signed-rank test cannot reach two-sided p < 0.05, so a
paired t-test is used.

```{r stats}
paired_ttest_p <- function(x, y) {
  keep <- complete.cases(x, y)
  x <- x[keep]
  y <- y[keep]
  if (length(x) < 2) return(NA_real_)
  tryCatch(t.test(x, y, paired = TRUE)$p.value, error = function(e) NA_real_)
}

p_to_stars <- function(p) {
  dplyr::case_when(is.na(p) ~ NA_character_,
                   p < 0.0001 ~ "****",
                   p < 0.001 ~ "***",
                   p < 0.01 ~ "**",
                   p < 0.05 ~ "*")
}

wt_rich <- raw_df %>%
  filter(strain == "WT") %>%
  select(dose, `repeat`, wt_viability = viability)

strain_stats <- raw_df %>%
  filter(strain != "WT", dose != "Untreated") %>%
  left_join(wt_rich, by = c("dose", "repeat")) %>%
  group_by(dose, strain) %>%
  summarise(n_pairs = sum(complete.cases(viability, wt_viability)),
            p_raw = paired_ttest_p(viability, wt_viability),
            .groups = "drop") %>%
  group_by(dose) %>%
  mutate(p_adj = p.adjust(p_raw, method = "holm"),
         stars = p_to_stars(p_adj)) %>%
  ungroup() %>%
  arrange(dose, strain)
write_csv(strain_stats, out_path("strain_vs_WT_paired_ttest.csv"))
knitr::kable(strain_stats, digits = 6)
```

Comparison with the saved results of the original analysis:

```{r verify}
orig_stats <- read_csv(data_path(paste0("strain_vs_WT_paired_ttest_", version_name, ".csv")),
                       col_types = cols(dose = col_character(), stars = col_character()),
                       show_col_types = FALSE)
cmp <- strain_stats %>%
  mutate(dose = as.character(dose), strain = as.character(strain)) %>%
  inner_join(orig_stats, by = c("dose", "strain"), suffix = c("", "_orig"))
data.frame(
  check = c("rows matched", "max |p_raw diff|", "max |p_adj diff|", "stars identical"),
  value = c(nrow(cmp),
            max(abs(cmp$p_raw - cmp$p_raw_orig)),
            max(abs(cmp$p_adj - cmp$p_adj_orig)),
            all(coalesce(cmp$stars, "") == coalesce(cmp$stars_orig, "")))
)
```

## Figure panel: viability across micafungin doses

```{r figure-micafungin, fig.width = 12, fig.height = 6}
star_offset <- 0.05 * max(raw_summary$mean_viability +
                            dplyr::coalesce(raw_summary$se_viability, 0), na.rm = TRUE)

# All strain x dose combinations are kept (untested ones get an empty label) so that
# position_dodge() aligns the labels with the 5 dodged bars per dose
raw_annot <- raw_summary %>%
  left_join(strain_stats %>% select(dose, strain, stars), by = c("dose", "strain")) %>%
  mutate(stars = tidyr::replace_na(stars, ""),
         y_star = mean_viability + dplyr::coalesce(se_viability, 0) + star_offset)

p_micafungin <- ggplot(raw_summary, aes(x = dose, y = mean_viability, fill = strain)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75,
           color = "black", linewidth = 0.25) +
  geom_errorbar(aes(ymin = pmax(mean_viability - dplyr::coalesce(se_viability, 0), 0),
                    ymax = pmin(mean_viability + dplyr::coalesce(se_viability, 0), 1)),
                position = position_dodge(width = 0.8), width = 0.2, linewidth = 0.4) +
  geom_point(data = raw_df, aes(x = dose, y = viability, group = strain),
             inherit.aes = FALSE, alpha = 0.3, size = 1.5,
             position = position_jitterdodge(jitter.width = 0.12, jitter.height = 0,
                                             dodge.width = 0.8)) +
  geom_text(data = raw_annot, aes(x = dose, y = y_star, group = strain, label = stars),
            inherit.aes = FALSE, position = position_dodge(width = 0.8), size = 3) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.25), labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Micafungin dose (ug/mL)", y = "Viability",
       title = "KO Strain Viability in Rich Media",
       subtitle = "Viability = Live_cells / (Live_cells + Dead_cells)",
       fill = "Strain",
       caption = paste0(
         "Bars = mean of triplicates; error bars = SE; points = individual replicates\n",
         "Stars: two-sided paired t-test, KO strain vs WT at each dose on raw viability ",
         "(paired by repeat), Holm-corrected per dose\n",
         "ns >= 0.05, * < 0.05, ** < 0.01, *** < 0.001, **** < 0.0001")) +
  cowplot::theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        legend.position = "right")
p_micafungin

ggsave(out_path("Figure_6_TFKO_micafungin_quiescent.pdf"), p_micafungin, width = 12, height = 6)
ggsave(out_path("Figure_6_TFKO_micafungin_quiescent.png"), p_micafungin, width = 12, height = 6,
       dpi = 300)
```

## Session info

```{r session-info}
sessionInfo()
```
