---
title: "Figure 6: Gene expression analysis of TF KOs"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
  padj_cutoff: 0.05
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)
suppressPackageStartupMessages({
  library(DESeq2)
  library(clusterProfiler)
  library(ggplot2)
  library(data.table)
})
set.seed(1)
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(f) file.path(params$data_dir, f)
out_path <- function(...) file.path(params$out_dir, ...)
```

## Figure legend

**Figure 6. Transcriptional regulation of quiescence.** Gene expression analysis of TF KOs (panel B in the legend; panel D in the current figure draft).

For each transcription factor knockout (BG, NDT, NRG, STP) we compare quiescent (72 h) with
proliferative (6 h) cultures in triplicate, using DESeq2 (`~ time`, Wald test, 72 h vs 6 h).
Genes with adjusted p-value < `r params$padj_cutoff` are called up- or down-regulated at 72 h.
Gene-set enrichment uses CGD GO annotations and CGD pathways, by over-representation analysis
(ORA) of the up- and down-regulated genes and by GSEA on the DESeq2 Wald statistic.

## Input data

All inputs are in `data/`:

| File | Description |
|---|---|
| `CandidaTFKO5_NovaSeq_24_gene_count_matrix_rounded.tsv` | Gene-level read counts (Assembly 22, haploid gene IDs), 24 libraries |
| `C_albicans_A22_r228_gene_info.tsv` | Gene ID, gene name, orf19 ID and description (CGD Assembly 22, r228) |
| `cgd_C_albicans_SC5314.gaf.gz` | CGD GO annotation file (GAF) |
| `pathwaysAndGenes.tab` | CGD biochemical pathways and their genes |

The raw reads are available from SRA (PRJNA1271226).

```{r load-counts}
counts <- fread(data_path("CandidaTFKO5_NovaSeq_24_gene_count_matrix_rounded.tsv"),
                data.table = FALSE, check.names = FALSE)
count_cols <- grep("^CandidaTFKO5_", names(counts), value = TRUE)
count_cols <- count_cols[!grepl("WT_72h", count_cols)]
stopifnot(length(count_cols) == 24L)
count_matrix <- as.matrix(counts[, count_cols, drop = FALSE])
rownames(count_matrix) <- counts$gene_id
storage.mode(count_matrix) <- "integer"

info <- fread(data_path("C_albicans_A22_r228_gene_info.tsv"), data.table = FALSE)
info <- info[!duplicated(info$gene_id), , drop = FALSE]
rownames(info) <- info$gene_id

data.frame(library = count_cols, total_counts = colSums(count_matrix))
```

## Gene-set annotation

GO terms (GAF qualifiers `involved_in`, `enables`, `located_in`) and CGD pathways are mapped
to Assembly 22 gene IDs through gene names and orf19 IDs.

```{r annotation}
key <- data.frame(gene_id = info$gene_id, gene_name = info$gene_name,
                  orf19 = info$orf19, stringsAsFactors = FALSE)
key_long <- unique(do.call(rbind, lapply(names(key)[-1], function(k) {
  data.frame(token = key[[k]], gene_id = key$gene_id, stringsAsFactors = FALSE)
})))
key_long <- key_long[!is.na(key_long$token) & nzchar(key_long$token), , drop = FALSE]

token_to_gene <- function(tokens) {
  tokens <- unique(unlist(strsplit(tokens, "\\|", fixed = FALSE)))
  out <- key_long$gene_id[match(tokens, key_long$token)]
  unique(out[!is.na(out)])
}

gaf <- fread(data_path("cgd_C_albicans_SC5314.gaf.gz"), sep = "\t", header = FALSE,
             comment.char = "!", data.table = FALSE, quote = "")
gaf <- gaf[!is.na(gaf$V5) & gaf$V4 %in% c("involved_in", "enables", "located_in"), ]
go_pairs <- unique(do.call(rbind, lapply(seq_len(nrow(gaf)), function(i) {
  ids <- token_to_gene(c(gaf$V3[i], gaf$V11[i]))
  if (!length(ids)) return(NULL)
  data.frame(term = gaf$V5[i], gene_id = ids, stringsAsFactors = FALSE)
})))

pathways <- fread(data_path("pathwaysAndGenes.tab"), sep = "\t", header = TRUE,
                  data.table = FALSE, quote = "", fill = TRUE)
path_pairs <- unique(do.call(rbind, lapply(seq_len(nrow(pathways)), function(i) {
  ids <- token_to_gene(pathways$`GENE-IDS`[i])
  if (!length(ids)) return(NULL)
  data.frame(term = pathways$NAME[i], gene_id = ids, stringsAsFactors = FALSE)
})))

term2gene <- unique(rbind(go_pairs, path_pairs))
term2gene <- term2gene[!is.na(term2gene$term) & nzchar(term2gene$term), ]
stopifnot(nrow(term2gene) > 0)
fwrite(term2gene, out_path("term2gene.tsv"), sep = "\t")
```

## Differential expression: 72 h vs 6 h in each TF KO

```{r deseq2}
conditions <- c("BG", "NDT", "NRG", "STP")
all_res <- list()
all_summary <- list()

for (condition in conditions) {
  samples <- grep(paste0("^CandidaTFKO5_", condition, "_(6h|72h)_"), count_cols, value = TRUE)
  time <- factor(ifelse(grepl("_72h_", samples), "72h", "6h"), levels = c("6h", "72h"))
  stopifnot(length(samples) == 6L, all(table(time) == 3L))

  dds <- DESeqDataSetFromMatrix(
    countData = count_matrix[, samples, drop = FALSE],
    colData = data.frame(row.names = samples, time = time),
    design = ~ time
  )
  dds <- DESeq(dds, quiet = TRUE)
  res <- as.data.frame(results(dds, contrast = c("time", "72h", "6h"),
                               independentFiltering = TRUE))
  res$gene_id <- rownames(res)
  res <- merge(res, info[, c("gene_id", "gene_name", "orf19", "note")],
               by = "gene_id", all.x = TRUE, sort = FALSE)
  res <- res[order(res$padj, res$pvalue, na.last = TRUE), ]
  sig <- !is.na(res$padj) & res$padj < params$padj_cutoff
  res$regulation <- ifelse(sig & res$log2FoldChange > 0, "up_72h",
                           ifelse(sig & res$log2FoldChange < 0, "down_72h", "not_significant"))

  condition_dir <- out_path(condition)
  dir.create(condition_dir, showWarnings = FALSE)
  fwrite(res, file.path(condition_dir, paste0(condition, "_DESeq2_72h_vs_6h.tsv")),
         sep = "\t", na = "NA")

  normalized <- as.data.frame(counts(dds, normalized = TRUE))
  normalized$gene_id <- rownames(normalized)
  normalized <- merge(normalized, info[, c("gene_id", "gene_name", "orf19")],
                      by = "gene_id", all.x = TRUE, sort = FALSE)
  fwrite(normalized, file.path(condition_dir, paste0(condition, "_normalized_counts.tsv")),
         sep = "\t", na = "NA")

  # GSEA on the Wald statistic. clusterProfiler >= 4.20 runs GSEA through enrichit and ignores
  # a `seed` argument, so the permutation-based p-values are only reproducible with set.seed().
  ranked <- res$stat
  names(ranked) <- res$gene_id
  ranked <- sort(ranked[is.finite(ranked)], decreasing = TRUE)
  set.seed(123)
  gsea <- GSEA(ranked, TERM2GENE = term2gene[, c("term", "gene_id")],
               minGSSize = 5, maxGSSize = 500, pvalueCutoff = 1,
               verbose = FALSE)
  gsea_df <- if (is.null(gsea)) data.frame() else as.data.frame(gsea)
  fwrite(gsea_df, file.path(condition_dir, paste0(condition, "_GSEA.tsv")),
         sep = "\t", na = "NA")

  # ORA of up- and down-regulated genes against all tested genes
  universe <- rownames(dds)
  for (direction in c("up_72h", "down_72h")) {
    selected <- res$gene_id[res$regulation == direction & res$gene_id %in% universe]
    ora <- enricher(selected, universe = universe,
                    TERM2GENE = term2gene[, c("term", "gene_id")],
                    minGSSize = 3, maxGSSize = 500,
                    pvalueCutoff = 1, qvalueCutoff = 1)
    ora_df <- if (is.null(ora)) data.frame() else as.data.frame(ora)
    fwrite(ora_df, file.path(condition_dir, paste0(condition, "_ORA_", direction, ".tsv")),
           sep = "\t", na = "NA")
  }

  all_res[[condition]] <- cbind(condition = condition, res)
  all_summary[[condition]] <- data.frame(
    condition = condition,
    tested_genes = nrow(res),
    significant_genes = sum(sig),
    up_72h = sum(res$regulation == "up_72h" & res$gene_id %in% universe),
    down_72h = sum(res$regulation == "down_72h" & res$gene_id %in% universe),
    gsea_terms = nrow(gsea_df)
  )
}

summary_df <- do.call(rbind, all_summary)
fwrite(summary_df, out_path("analysis_summary.tsv"), sep = "\t")
knitr::kable(summary_df, row.names = FALSE)
```

## Figure panel: volcano plots

Volcano plots of quiescent (72 h) vs proliferative (6 h) gene expression for each TF KO.

```{r figure-volcano, fig.width = 10, fig.height = 8}
# Genes without an adjusted p-value (low counts or outliers) cannot be placed on the y-axis
res_all <- do.call(rbind, all_res)
res_all <- res_all[!is.na(res_all$padj), ]
res_all$minus_log10_padj <- -log10(pmax(res_all$padj, .Machine$double.xmin))
facet_labels <- setNames(
  sprintf("%s (up: %d, down: %d)", summary_df$condition, summary_df$up_72h, summary_df$down_72h),
  summary_df$condition
)
res_all$condition <- factor(facet_labels[res_all$condition], levels = facet_labels[conditions])

p_rnaseq <- ggplot(res_all, aes(x = log2FoldChange, y = minus_log10_padj, color = regulation)) +
  geom_point(alpha = 0.65, size = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3) +
  geom_hline(yintercept = -log10(params$padj_cutoff), linetype = "dashed", linewidth = 0.3) +
  facet_wrap(~ condition, ncol = 2) +
  scale_color_manual(values = c(up_72h = "#B2182B", down_72h = "#2166AC",
                                not_significant = "grey70"),
                     labels = c(up_72h = "Up in quiescence", down_72h = "Down in quiescence",
                                not_significant = "Not significant")) +
  labs(x = expression(log[2]~fold~change~(72~h~"/"~6~h)),
       y = expression(-log[10]~adjusted~p-value), color = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "top", strip.background = element_blank(),
        strip.text = element_text(face = "bold"))
p_rnaseq

ggsave(out_path("Figure_6_TFKO_RNAseq_volcano.pdf"), p_rnaseq, width = 10, height = 8)
ggsave(out_path("Figure_6_TFKO_RNAseq_volcano.png"), p_rnaseq, width = 10, height = 8, dpi = 300)
```

## Session info

```{r session-info}
sessionInfo()
```
