---
title: "Figure 6: Viability screen of the TF KO collection in quiescence"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
  version_name: "OzanImir_20260310_v1"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(tidyverse)
})
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(...) file.path(params$data_dir, "flow_screen", ...)
out_path <- function(...) file.path(params$out_dir, ...)
version_name <- params$version_name
results_file <- data_path(paste0(version_name, "_viability_results.csv"))
has_results <- file.exists(results_file)
```

## Figure legend

**Figure 6. Transcriptional regulation of quiescence.** A) Screen of TF KOs for decreased survival
in quiescence (panel A in the current figure draft).

Each strain of the transcription factor knockout collection (166 KOs) and the wild-type control
SC5314 was grown to quiescence for 3 days, stained with SYTO9 and propidium iodide (PI) and
measured on a Cytek flow cytometer. Viability = Live cells / (Live cells + Dead cells) x 100.
Strains are ranked by viability; SC5314 is shown in red with a dashed reference line.

## Input data

All inputs are in `data/flow_screen/`:

| File | Description |
|---|---|
| `sample_sheet_vignette_simple.csv` | FCS file name to strain ID (Strain001-Strain166, SC5314) |
| `Vignette-Experiment-Details.csv` | CytoExploreR experiment details (same content as the sample sheet) |
| `Vignette-Experiment-Markers.csv` | CytoExploreR channel/marker table (SYTO9 B7-A, PI YG4-A) |
| `cytek_gating_OzanImir_20260310_v1.csv` | CytoExploreR gating template (Cells, Single_cells, Live_cells, Dead_cells) |
| `OzanImir_20260310_v1_viability_results.csv` | Per-strain gated counts and viability. **Not available yet** (to be added) |

The raw FCS files are not distributed. The per-strain summary table written by the gating step
(`OzanImir_20260310_v1_viability_results.csv`) was not kept, so the plot chunk below only runs once
that table is added to `data/flow_screen/`.

Strain IDs map to plate wells; the strain-to-gene plate map is to be added.

## Gating (not run: requires FCS files)

Gating was done with CytoExploreR on the FCS files (not distributed), using the gating template in
`data/flow_screen/`. Transforms: arcsinh, cofactor 150 for SYTO9 (B7-A) and PI (YG4-A), cofactor
500 for FSC/SSC. Gates: Cells (FSC-A/SSC-A) > Single_cells (SSC-A/SSC-H) > Live_cells and
Dead_cells (YG4-A/B7-A).

```{r gating, eval = FALSE}
library(CytoExploreR)
library(flowWorkspace)

# FCS files are expected in data/flow_screen/FCS_files (not distributed)
my_gating_set <- cyto_setup(path = data_path("FCS_files"), restrict = TRUE, select = "fcs")

# Attach the sample metadata (Vignette-Experiment-Details.csv = sample sheet joined to file names)
experiment_details <- read_csv(data_path("Vignette-Experiment-Details.csv"), show_col_types = FALSE)
for (i in seq_along(experiment_details)) {
  pData(my_gating_set)[names(experiment_details)[i]] <- experiment_details[[i]]
}

SYTO9_trans <- cyto_transformer_arcsinh(my_gating_set, channels = c("B7-A"), cofactor = 150)
PI_trans <- cyto_transformer_arcsinh(my_gating_set, channels = c("YG4-A"), cofactor = 150)
FSC_SSC_trans <- cyto_transformer_arcsinh(my_gating_set,
                                          channels = c("FSC-A", "FSC-H", "SSC-A", "SSC-H"),
                                          cofactor = 500)
combined_trans <- cyto_transformer_combine(SYTO9_trans, PI_trans, FSC_SSC_trans)
transformed_gating_set <- cyto_transform(my_gating_set, trans = combined_trans)

cyto_gatingTemplate_apply(transformed_gating_set,
                          gatingTemplate = data_path(paste0("cytek_gating_", version_name, ".csv")))

# Raw cell counts for Single_cells, Live_cells and Dead_cells
pop_counts <- gs_pop_get_stats(transformed_gating_set,
                               nodes = c("Single_cells", "Live_cells", "Dead_cells"),
                               type = "count")
pop_counts_wide <- as_tibble(pop_counts) %>%
  pivot_wider(names_from = pop, values_from = count)

metadata <- pData(transformed_gating_set) %>%
  as_tibble(rownames = "sample") %>%
  select(sample, strain)

pop_counts_wide %>%
  left_join(metadata, by = "sample") %>%
  mutate(viability = Live_cells / (Live_cells + Dead_cells) * 100) %>%
  select(sample, strain, Single_cells, Live_cells, Dead_cells, viability) %>%
  write_csv(results_file)
```

## Viability ranking

```{r viability, eval = has_results}
viability_data <- read_csv(results_file, show_col_types = FALSE) %>%
  mutate(is_control = strain == "SC5314") %>%
  arrange(viability) %>%
  mutate(strain = factor(strain, levels = unique(strain)))
stopifnot(nrow(viability_data) == 167L)

sc5314_viability <- viability_data %>%
  filter(is_control) %>%
  pull(viability)

viability_data %>%
  select(strain, Single_cells, Live_cells, Dead_cells, viability) %>%
  knitr::kable(digits = 2)
```

```{r no-results, echo = FALSE, eval = !has_results, results = "asis"}
cat("**Summary table not available:** `", basename(results_file),
    "` is not in `data/flow_screen/`, so the panel cannot be drawn yet.\n", sep = "")
```

## Figure panel: ranked viability of the TF KO collection

```{r figure-screen, eval = has_results, fig.width = 22, fig.height = 6}
p_screen <- ggplot(viability_data, aes(x = strain, y = viability, fill = is_control)) +
  geom_col(width = 0.8) +
  geom_hline(yintercept = sc5314_viability, linetype = "dashed",
             color = "firebrick", linewidth = 0.7) +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
                    labels = c("FALSE" = "TF KO Strain", "TRUE" = "SC5314 (WT Control)"),
                    name = NULL) +
  scale_y_continuous(limits = c(0, 105), breaks = seq(0, 100, 10), expand = c(0, 0)) +
  labs(title = "KO Strain Collection Viability After 3 Day Incubation",
       subtitle = "Viability = Live cells / (Live + Dead cells) × 100",
       x = "Strain", y = "Viability (%)",
       caption = paste0("Red dashed line: SC5314 (WT) viability = ",
                        round(sc5314_viability, 1), "%\n",
                        "n = ", nrow(viability_data), " strains, sorted by viability (ascending)")) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 5),
        plot.title = element_text(face = "bold", size = 12),
        plot.caption = element_text(hjust = 0),
        legend.position = "top")
p_screen

ggsave(out_path("Figure_6_TFKO_viability_screen.pdf"), p_screen, width = 22, height = 6)
ggsave(out_path("Figure_6_TFKO_viability_screen.png"), p_screen, width = 22, height = 6, dpi = 300)
```

## Session info

```{r session-info}
sessionInfo()
```
