---
title: "Figure 5: Antifungal response of 20 isolates and SC5314, proliferative vs quiescent"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6)
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(ggplot2)
})
set.seed(1)
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(f) file.path(params$data_dir, f)
out_path <- function(...) file.path(params$out_dir, ...)
```

## Figure legend

**Figure 5. Quiescent cells exhibit increased survival when treated with antifungals.**
C) Viability for proliferative and quiescent cells using 20 genetically diverse strains and the lab
strain SC5314 at the LD50 drug concentration identified for proliferative SC5314 cells. Strains were
treated with either 1µg/mL caspofungin, 1µg/mL micafungin, or 20µg/mL amphotericin B. Black line
inside box shows median viability of 21 strains; box range indicates interquartile range (IQR);
whiskers indicate 1.5 times the IQR from median. Viability of each strain is indicated as a black
dot; SC5314 viability is indicated with a yellow dot; the p-value was determined by Student's t-test,
∗p < 0.05, ∗∗p < 0.01, ∗∗∗p < 0.001.

The code below uses paired t-tests (paired by strain) and plots SC5314 in orange.

## Input data

All inputs are in `data/`. Each row is one well with gated Live/Dead percentages (raw FCS files are
not distributed). `Dose` is coded 0 = untreated control and 1 = treated (1 µg/mL caspofungin,
1 µg/mL micafungin, or the amphotericin B treatment dose).

| File | Description |
|---|---|
| `Isolates_AmpB_Exp.csv` | Amphotericin B, proliferative cultures, 20 isolates + SC5314, control and treated |
| `Isolates_AmpB_Qui.csv` | Amphotericin B, quiescent cultures |
| `Isolates_Caspofungin_Exp.csv` | Caspofungin, proliferative cultures |
| `Isolates_Caspofungin_Qui.csv` | Caspofungin, quiescent cultures |
| `Isolates_Micafungin_Exp.csv` | Micafungin, proliferative cultures |
| `Isolates_Micafungin_Qui.csv` | Micafungin, quiescent cultures |

```{r load-data}
read_data <- function(file) {
  read.table(file = data_path(file), header = TRUE, sep = ",", dec = ".", fill = TRUE,
             fileEncoding = "UTF-8-BOM")
}
AmpB_Exp <- read_data("Isolates_AmpB_Exp.csv")
AmpB_Qui <- read_data("Isolates_AmpB_Qui.csv")
Casp_Exp <- read_data("Isolates_Caspofungin_Exp.csv")
Casp_Qui <- read_data("Isolates_Caspofungin_Qui.csv")
Mica_Exp <- read_data("Isolates_Micafungin_Exp.csv")
Mica_Qui <- read_data("Isolates_Micafungin_Qui.csv")

df <- rbind(AmpB_Exp, AmpB_Qui, Casp_Exp, Casp_Qui, Mica_Exp, Mica_Qui)
count(df, Treatment, Growth, Dose)
```

## Viability per strain and condition

Viability = Live% / (Live% + Dead%). Dose 0 wells from all three drug files are pooled as
"Untreated"; Dose 1 wells keep the drug name.

```{r prepare}
df_summary <- df %>%
  group_by(Sample, Growth, Treatment, Dose) %>%
  summarise(
    Total = Live_percent + Dead_percent,
    Live_percent = Live_percent / Total,
    .groups = "drop"
  )

plot_data <- df_summary %>%
  mutate(Condition = case_when(
    Dose == 0 ~ "Untreated",
    TRUE ~ Treatment
  )) %>%
  filter(Dose == 0 | (Dose == 1 & Treatment %in% c("Caspofungin", "Micafungin", "AmphotericinB")))

level_order <- c(
  "Untreated Exponential", "Untreated Quiescent",
  "Caspofungin Exponential", "Caspofungin Quiescent",
  "Micafungin Exponential", "Micafungin Quiescent",
  "AmphotericinB Exponential", "AmphotericinB Quiescent"
)

plot_data <- plot_data %>%
  mutate(Group_Label = paste(Condition, Growth, sep = " "),
         Group_Label = factor(Group_Label, levels = level_order))

plot_data %>%
  group_by(Group_Label) %>%
  summarise(n = n(), median_viability = median(Live_percent), .groups = "drop")
```

## Paired t-tests, proliferative vs quiescent

```{r paired-tests}
fmt <- function(p) {
  if (is.na(p)) return(list(p_label = "n/a", stars = "n.s."))
  if (p < 0.001) return(list(p_label = "p < 0.001", stars = "***"))
  if (p < 0.01) return(list(p_label = sprintf("p = %.3f", p), stars = "**"))
  if (p < 0.05) return(list(p_label = sprintf("p = %.3f", p), stars = "*"))
  list(p_label = sprintf("p = %.3f", p), stars = "n.s.")
}

paired_tests <- plot_data %>%
  group_by(Condition) %>%
  summarise(
    data_wide = list(
      pivot_wider(
        pick(everything()),
        id_cols = Sample,
        names_from = Growth,
        values_from = Live_percent
      )
    ),
    .groups = "drop"
  ) %>%
  mutate(
    test_results = map(data_wide, ~{
      d <- .x
      if (!all(c("Exponential", "Quiescent") %in% names(d))) {
        return(list(p = NA_real_, n = 0))
      }
      d_clean <- d %>% filter(!is.na(Exponential), !is.na(Quiescent))
      n_pairs <- nrow(d_clean)
      if (n_pairs >= 2) {
        p_val <- tryCatch(
          t.test(d_clean$Exponential, d_clean$Quiescent, paired = TRUE)$p.value,
          error = function(e) NA_real_
        )
      } else {
        p_val <- NA_real_
      }
      list(p = p_val, n = n_pairs)
    }),
    p_value = map_dbl(test_results, "p"),
    n_pairs = map_int(test_results, "n"),
    stats = map(p_value, fmt),
    p_label = map_chr(stats, "p_label"),
    stars = map_chr(stats, "stars")
  ) %>%
  select(Condition, p_value, n_pairs, p_label, stars)

paired_tests
write.csv(paired_tests, out_path("Figure_5_paired_tests.csv"), row.names = FALSE)
```

For "Untreated", every strain has three Dose 0 values per growth phase (one per drug file), so
`pivot_wider()` returns list-columns and the paired test returns `NA`.

## Plot

```{r plot-drug-response, fig.width = 10, fig.height = 6}
p <- ggplot(plot_data, aes(x = Group_Label, y = Live_percent, fill = Growth)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(data = subset(plot_data, !Sample %in% c("SC5314")),
              position = position_jitter(width = 0.08, height = 0),
              size = 1, color = "black", alpha = 0.7) +
  geom_jitter(data = subset(plot_data, Sample %in% c("SC5314")),
              position = position_jitter(width = 0.08, height = 0),
              size = 2, color = "orange") +
  scale_y_continuous(name = "Viability (%)", limits = c(0, 1)) +
  scale_fill_manual(values = c("Exponential" = "#F8766D", "Quiescent" = "#00BFC4")) +
  theme_gray() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_blank(),
    legend.position = "right",
    panel.grid.major.x = element_blank()
  ) +
  labs(fill = "Growth Phase",
       title = "Drug Response: Exponential vs Quiescent",
       subtitle = "Paired t-tests comparing growth phases (Dose 0 = Untreated, Dose 1 = Treated)")
p

ggsave(out_path("drug_response_4pairs.png"), p, width = 10, height = 6, dpi = 300)
ggsave(out_path("drug_response_4pairs.pdf"), p, width = 10, height = 6)
```

## Session info

```{r session-info}
sessionInfo()
```
