---
title: "Figure 5: Micafungin dose response of clinical isolates p37005 and p57055"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 7)
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
})
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(f) file.path(params$data_dir, f)
out_path <- function(...) file.path(params$out_dir, ...)
```

## Figure legend

**Figure 5. Quiescent cells exhibit increased survival when treated with antifungals.**
B) Viability of proliferative and quiescent cells using clinically isolates p57055 and p37005 at
of 0.001 µg/mL to 10 µg/mL micafungin.

Viability is the fraction of PI/SYTO9-gated events that are live, Live% / (Live% + Dead%), for each
well of the 20-isolate micafungin plates (one well per isolate and dose). In the screen,
p37005 is `Sample1` (solid line) and p57055 is `Sample17` (dashed line).

## Input data

All inputs are in `data/`. Each row is one well (gated Live/Dead percentages exported from the
Cytek gating of the full screen; raw FCS files are not distributed).

| File | Description |
|---|---|
| `Isolates_Micafungin_Exponential_1-10.CSV` | Proliferative cultures, isolates Sample1–Sample10, micafungin 0 (Blank) to 10 µg/mL |
| `Isolates_Micafungin_Exponential_11-20.CSV` | Proliferative cultures, isolates Sample11–Sample20 |
| `Isolates_Micafungin_Quiescent_1-10.CSV` | Quiescent cultures, isolates Sample1–Sample10 |
| `Isolates_Micafungin_Quiescent_11-20.CSV` | Quiescent cultures, isolates Sample11–Sample20 |

## Data preparation

```{r load-data}
files <- c("Isolates_Micafungin_Exponential_1-10.CSV", "Isolates_Micafungin_Exponential_11-20.CSV",
           "Isolates_Micafungin_Quiescent_1-10.CSV", "Isolates_Micafungin_Quiescent_11-20.CSV")
combined_data_clean <- do.call(rbind, lapply(data_path(files), read.csv,
                                             fileEncoding = "UTF-8-BOM"))

# Viability = Live / (Live + Dead)
combined_data_clean$Live_Dead_ratio <- with(combined_data_clean,
                                            Live_percent / (Live_percent + Dead_percent))

table(combined_data_clean$Growth, combined_data_clean$Dose)
```

## Micafungin response of p37005 (Sample1) and p57055 (Sample17)

```{r subset}
subset_data <- subset(combined_data_clean,
                      Sample %in% c("Sample1", "Sample17") & Treatment == "Micafungin")

# Untreated wells (Dose 0) fall outside the log10 axis and are not plotted
subset_data <- subset(subset_data, Dose > 0)

subset_data %>%
  arrange(Sample, Growth, Dose) %>%
  select(Sample, Growth, Dose, Live_percent, Dead_percent, Live_Dead_ratio)
```

```{r plot-micafungin, fig.width = 7, fig.height = 7}
custom_labels <- c("-3", "-2", "-1", "0", "1")
custom_breaks <- c(0.001, 0.01, 0.1, 1, 10)

p <- ggplot(subset_data, aes(x = Dose, y = Live_Dead_ratio, color = Growth, linetype = Sample)) +
  geom_line() +
  geom_point(size = 3) +
  scale_x_log10(labels = custom_labels, breaks = custom_breaks, limits = c(0.00001, 11)) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_linetype_manual(values = c("Sample1" = "solid", "Sample17" = "dashed"),
                        labels = c("Sample1" = "p37005 (Sample1)", "Sample17" = "p57055 (Sample17)")) +
  labs(x = "log10(Drug Concentration)(ug/mL)",
       y = "Viability") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p

ggsave(out_path("Figure_5_micafungin_p37005_p57055.png"), p, width = 7, height = 7, dpi = 300)
ggsave(out_path("Figure_5_micafungin_p37005_p57055.pdf"), p, width = 7, height = 7)
```

## Session info

```{r session-info}
sessionInfo()
```
