---
title: "Supplementary Figure 23: Viability screen of the TF KO collection in quiescence"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
  version_name: "OzanImir_20260310_v1"
  fcs_root: "/scratch/obi203/data/Cytek Flow Cytometry"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(tidyverse)
})
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(...) file.path(params$data_dir, ...)
out_path <- function(...) file.path(params$out_dir, ...)
version_name <- params$version_name
results_file <- data_path(paste0(version_name, "_viability_results.csv"))
fcs_dir <- file.path(params$fcs_root, "Knockout_Quiescence_Viability_2", "FCS_files")
fcs_available <- dir.exists(fcs_dir)
```

## Figure legend

**Supplementary Figure 23. Viability screen of the transcription factor knockout collection in
quiescence.** Each strain of the TF KO collection (166 KOs) and the wild-type control (WT, SC5314)
was grown to quiescence for 3 days in rich media, stained with SYTO9 and PI and measured by flow
cytometry (one well per strain). Viability = Live / (Live + Dead) × 100. Strains are ranked by
viability. WT is shown in red, and the dashed line marks WT viability (98.7%). The shaded region marks the
19 strains with the lowest viability.

(Full legend text is in the folder README.)

## Input data

| File | Description |
|---|---|
| `data/sample_sheet_vignette_simple.csv` | FCS file name to strain ID (Strain001-Strain166, SC5314) |
| `data/Vignette-Experiment-Markers.csv` | CytoExploreR channel/marker table (SYTO9 B7-A, PI YG4-A) |
| `data/cytek_gating_OzanImir_20260310_v1.csv` | CytoExploreR gating template (Cells, Single_cells, Live_cells, Dead_cells) |
| `data/OzanImir_20260310_v1_viability_results.csv` | Per-strain gated counts and viability (167 strains), regenerated from the FCS files by the gating chunk below |
| `data/OzanImir_20260310_v1_viability_from_plot.csv` | Viability values (2 decimals) read from the original plot `OzanImir_20260310_v1_viability_plot.pdf` (used for verification only) |

Strain IDs (Strain001-Strain166) are plate positions. The strain-to-gene plate map is not
available, so strains are labelled by plate ID.

## Gating (runs only when the FCS files are available)

The raw FCS files are not distributed. They are read from `params$fcs_root`. The gating uses the FCS
reader and gate code from Supplementary Figure 22 (no CytoExploreR):

* `cyto_transformer_arcsinh()` in the source notebook applied the flowJo arcsinh with its defaults
  (T = 262144, M = 4.5, A = 0). The `cofactor` arguments had no effect.
* Gates: Cells (FSC-A/SSC-A) > Single_cells (SSC-A/SSC-H) > Live_cells and Dead_cells (YG4-A = PI vs
  B7-A = SYTO9).

```{r functions}
# Minimal FCS 3.x reader (float32 list-mode data, as written by the Cytek Aurora)
read_fcs <- function(path) {
  con <- file(path, "rb"); on.exit(close(con))
  hdr <- readChar(con, 58, useBytes = TRUE)
  ts <- as.integer(substr(hdr, 11, 18)); te <- as.integer(substr(hdr, 19, 26))
  seek(con, ts); txt <- readChar(con, te - ts + 1, useBytes = TRUE)
  d <- substr(txt, 1, 1); parts <- strsplit(substring(txt, 2), d, fixed = TRUE)[[1]]
  kw <- setNames(parts[seq(2, length(parts), 2)], toupper(parts[seq(1, length(parts), 2)]))
  stopifnot(kw[["$DATATYPE"]] == "F")
  npar <- as.integer(kw[["$PAR"]]); ntot <- as.integer(kw[["$TOT"]])
  endian <- if (kw[["$BYTEORD"]] == "1,2,3,4") "little" else "big"
  seek(con, as.numeric(trimws(kw[["$BEGINDATA"]])))
  m <- matrix(readBin(con, "numeric", n = npar * ntot, size = 4, endian = endian), ncol = npar, byrow = TRUE)
  colnames(m) <- vapply(seq_len(npar), function(i) kw[[paste0("$P", i, "N")]], "")
  m
}

# flowJo arcsinh as used by CytoExploreR (T = 262144, M = 4.5, A = 0)
fasinh <- function(x, t = 262144, m = 4.5) asinh(x * sinh(m * log(10)) / t) / (m * log(10))

# Polygon gates from a CytoExploreR gatingTemplate
parse_template <- function(csv) {
  gt <- read.csv(csv, stringsAsFactors = FALSE)
  setNames(lapply(seq_len(nrow(gt)), function(i) {
    nums <- as.numeric(strsplit(sub(".*boundaries = structure\\(c\\(([^)]*)\\).*", "\\1",
                                    gt$gating_args[i]), ",")[[1]])
    dims <- strsplit(gt$dims[i], ",")[[1]]
    list(parent = gt$parent[i], dims = dims, poly = matrix(nums, ncol = 2, dimnames = list(NULL, dims)))
  }), gt$alias)
}

apply_gates <- function(tm, gates) {
  keep <- list(root = rep(TRUE, nrow(tm)))
  for (g in names(gates)) {
    G <- gates[[g]]
    inside <- sp::point.in.polygon(tm[, G$dims[1]], tm[, G$dims[2]], G$poly[, 1], G$poly[, 2]) > 0
    keep[[g]] <- keep[[G$parent]] & inside
  }
  keep
}
```

```{r gating, eval = fcs_available}
sample_sheet <- read.csv(data_path("sample_sheet_vignette_simple.csv"))
gates <- parse_template(data_path(paste0("cytek_gating_", version_name, ".csv")))

gated <- bind_rows(lapply(seq_len(nrow(sample_sheet)), function(i) {
  raw <- read_fcs(file.path(fcs_dir, sample_sheet$name[i]))
  tm <- raw
  for (ch in c("FSC-A", "FSC-H", "SSC-A", "SSC-H", "YG4-A", "B7-A")) tm[, ch] <- fasinh(raw[, ch])
  k <- apply_gates(tm, gates)
  tibble(sample = sample_sheet$name[i], strain = sample_sheet$strain[i],
         Single_cells = sum(k$Single_cells), Live_cells = sum(k$Live_cells), Dead_cells = sum(k$Dead_cells))
})) %>%
  mutate(viability = Live_cells / (Live_cells + Dead_cells) * 100)

# Write the table on first run; afterwards the gating must reproduce it exactly
if (!file.exists(results_file)) write_csv(gated, results_file)
saved <- read_csv(results_file, show_col_types = FALSE)
stopifnot(identical(saved$strain, gated$strain),
          all(saved$Live_cells == gated$Live_cells), all(saved$Dead_cells == gated$Dead_cells))
```

## Viability ranking

```{r viability}
viability_data <- read_csv(results_file, show_col_types = FALSE) %>%
  mutate(is_control = strain == "SC5314") %>%
  arrange(viability) %>%
  mutate(rank = row_number(), strain = factor(strain, levels = strain))
stopifnot(nrow(viability_data) == 167L)
sc5314_viability <- viability_data$viability[viability_data$is_control]

viability_data %>%
  select(rank, strain, Single_cells, Live_cells, Dead_cells, viability) %>%
  head(25) %>%
  knitr::kable(digits = 2)
```

Comparison with the values in the original plot. The plot values were read to 2 decimals, so the
differences should be below 0.01 percentage points, and the ranking should be identical:

```{r verify}
from_plot <- read_csv(data_path(paste0(version_name, "_viability_from_plot.csv")), show_col_types = FALSE)
cmp <- inner_join(viability_data, from_plot, by = c("strain"))
data.frame(check = c("strains matched", "max |viability diff| (% points)", "ranking identical",
                     "SC5314 viability (%)"),
           value = c(nrow(cmp), signif(max(abs(cmp$viability - cmp$viability_pct)), 2),
                     identical(as.character(viability_data$strain), from_plot$strain),
                     round(sc5314_viability, 2)))
```

Strain125, the lowest-ranked strain, has only `r viability_data$Single_cells[1]` single cells,
compared with a median of `r format(median(viability_data$Single_cells), big.mark = ",")` across the screen.

## Figure

```{r figure, fig.width = 12, fig.height = 4.5}
ink <- "#2b2b2a"; ink2 <- "#6b6a65"
n_low <- 19  # shaded: the 19 lowest-viability strains

p_screen <- ggplot(viability_data, aes(strain, viability)) +
  annotate("rect", xmin = 0.4, xmax = n_low + 0.5, ymin = 0, ymax = 104,
           fill = "#1baf7a", alpha = 0.15) +
  geom_col(aes(fill = is_control), width = 0.8) +
  geom_hline(yintercept = sc5314_viability, linetype = "dashed", colour = "#c0392b", linewidth = 0.35) +
  scale_fill_manual(values = c(`FALSE` = "#5b87b8", `TRUE` = "#c0392b"), guide = "none") +
  scale_x_discrete(labels = function(x) ifelse(x == "SC5314", "WT", sub("Strain", "", x))) +
  scale_y_continuous(limits = c(0, 104), breaks = seq(0, 100, 20), expand = c(0, 0)) +
  labs(x = "TF KO strain (plate ID), ranked by viability", y = "Viability (%)") +
  theme_classic(base_size = 8) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 4.2, colour = ink2,
                                   face = ifelse(levels(viability_data$strain) == "SC5314", "bold", "plain")),
        axis.text.y = element_text(size = 10, colour = ink2),
        axis.title = element_text(size = 12, colour = ink),
        axis.line = element_line(colour = "grey55", linewidth = 0.3),
        axis.ticks = element_line(colour = "grey55", linewidth = 0.3),
        axis.ticks.length.x = unit(0.5, "mm"))
p_screen

ggsave(out_path("Supplementary_Figure_23.png"), p_screen, width = 12, height = 4.5, dpi = 300, bg = "white")
ggsave(out_path("Supplementary_Figure_23.pdf"), p_screen, width = 12, height = 4.5, bg = "white")
```

## Session info

```{r session-info}
sessionInfo()
```
