---
title: "Figure 4: Gene expression profiling in proliferative and quiescent cells"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
  n_clusters: 8
  rank_medium: "Rich"      # medium used to pick the top clusters and top genes
  rank_time: 72            # time point compared with 2 h to pick the top clusters and top genes
  n_top_clusters: 3
  n_top_genes: 10
  lfc_limit: 5             # colour scale limit for the panel B heatmap
  autophagy_terms: !r c("GO:0006914", "GO:0016236", "GO:0000045", "GO:0000422", "GO:0000423")  # panel D GO terms
  autophagy_highlight: "GO:0000045"  # autophagosome assembly; highlighted in panel D from late_time on
  late_time: 24            # first time point treated as quiescent (shaded in panels C and D)
  autophagy_genes: !r c("AUT7", "ATG1", "APG7", "ATG9", "ATG15", "PRB1")  # panel D trajectories
  highlight_gene: "AUT7"   # previously reported candidate (Atg8 orthologue), stained in this study
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(ggrepel)
  library(ggtext)
  library(ComplexHeatmap)
  library(circlize)
  library(patchwork)
  library(png)
  library(grid)
})
set.seed(1)
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(f) file.path(params$data_dir, f)
out_path <- function(...) file.path(params$out_dir, ...)
```

## Figure legend

**Figure 4. Gene expression profiling in proliferative and quiescent *Candida albicans* cells.**
A) Experimental design, sequencing pipeline, and computational strategy employed to quantify
transcript level changes across time. B) Hierarchical clustering defined 8 clusters that are
annotated using the most significant over-represented GO terms (p < 0.05). C) Expression of the
10 most strongly induced genes (`r params$rank_medium` `r params$rank_time` h vs 2 h) in each of
the three clusters with the highest induction in quiescence. D) Autophagy is induced
on entry into quiescence. Left, GO over-representation of autophagy terms among genes up-regulated at
each time point vs 2 h (padj < 0.05, log2FC ≥ 1) in rich and minimal medium; open grey circles, not
significant. Autophagosome assembly is highlighted from `r params$late_time` h on. Right, log2 fold change
vs 2 h of core autophagy genes in rich medium; *AUT7* (Atg8), a previously reported candidate that was also examined
by staining, is highlighted.

## Input data

All inputs are in `data/`. Except for the panel A schematic, they are outputs of
`CandidaTFKO5_WT_72h/analysis/WT_Rich_vs_Minimal_timecourse_DESeq2_ORA.Rmd`
(`new_72h_mode = "add"`, `n_clusters = 8`, GO BP).

| File | Description |
|---|---|
| `Figure_4A_schematic.png` | Panel A schematic, cropped from the original-submission `Figure_4.png` |
| `vst_batch_corrected.csv.gz` | DESeq2 VST, sequencing batch removed, 45 libraries (WT, rich and minimal, 2–72 h) |
| `timecourse_gene_clusters.csv` | Cluster assignment of the time-responsive genes (LRT padj < 0.05 in either medium) |
| `DESeq2_time_vs_2h.csv.gz` | Each time point vs 2 h within each medium; ashr-shrunken log2FC and padj |
| `ORA_GO_timecourse_clusters.csv` | GO BP over-representation per cluster (BH-adjusted p < 0.05) |
| `ORA_GO_contrasts_time_vs_2h.csv` | GO BP over-representation of the genes up or down at each time point vs 2 h within each medium (BH-adjusted p < 0.05) |

The raw reads are available from SRA (PRJNA1271226).

```{r load}
vst <- fread(data_path("vst_batch_corrected.csv.gz"), data.table = FALSE)
vmat <- as.matrix(vst[, -1]); rownames(vmat) <- vst[[1]]
clusters <- fread(data_path("timecourse_gene_clusters.csv"), data.table = FALSE)
de <- fread(data_path("DESeq2_time_vs_2h.csv.gz"))
ora <- fread(data_path("ORA_GO_timecourse_clusters.csv"))
ora_time <- fread(data_path("ORA_GO_contrasts_time_vs_2h.csv"))

# Sample groups. CandidaTFKO5_WT_72h-* are the Rich 72 h libraries from run HJT3JDRX7.
sample_medium <- ifelse(grepl("^WT_C_", colnames(vmat)), "Minimal", "Rich")
sample_time <- ifelse(grepl("^CandidaTFKO5_WT_72h", colnames(vmat)), 72L,
                      as.integer(sub("^WT_[YC]_(\\d+)-\\d$", "\\1", colnames(vmat))))
times <- sort(unique(sample_time))
media <- c("Rich", "Minimal")
group_levels <- paste(rep(media, each = length(times)), times, sep = "_")
sample_group <- factor(paste(sample_medium, sample_time, sep = "_"), levels = group_levels)
table(sample_medium, sample_time)
```

## Clusters and their GO annotation

The clustering is rebuilt here so that panel B can show the dendrogram. As in the analysis, VST values
of the time-responsive genes are averaged per medium × time group, z-scored across all 14 groups and
clustered with Ward.D2. Clusters are numbered by the time of their mean peak in rich medium. The check
below confirms the result is identical to the saved cluster assignment.

```{r clusters}
genes <- clusters$gene
gmeans <- sapply(group_levels, function(g) rowMeans(vmat[genes, sample_group == g, drop = FALSE]))
z <- t(scale(t(gmeans)))
z <- z[complete.cases(z), ]
hc <- hclust(dist(z), method = "ward.D2")
cl <- cutree(hc, k = params$n_clusters)
peak <- tapply(seq_along(cl), cl, function(i) which.max(colMeans(z[i, grep("^Rich", colnames(z)), drop = FALSE])))
cl <- match(cl, order(peak)); names(cl) <- rownames(z)
stopifnot(identical(unname(cl[clusters$gene]), clusters$cluster))
table(cluster = cl)
```

Each cluster is labelled with a short summary of its most significant GO terms. The top terms are
listed below so the labels can be checked against the ORA results.

```{r cluster-labels}
ora[, cluster := as.integer(sub("^C", "", Cluster))]
ora[order(cluster, p.adjust), head(.SD, 4), by = cluster][
  , .(cluster, Description, Count, FoldEnrichment = round(FoldEnrichment, 1), p.adjust = signif(p.adjust, 2))]

cluster_annotation <- c(
  "1" = "Translation & amino acid biosynthesis",
  "2" = "DNA replication & cell cycle",
  "3" = "Ribosome biogenesis & rRNA processing",
  "4" = "Golgi vesicle transport & ATP synthesis",
  "5" = "Small molecule & organic acid catabolism",
  "6" = "Peroxisome & vacuole organization",
  "7" = "Translational elongation & pheromone response",
  "8" = "Transport, cell adhesion & biofilm formation"
)
cluster_cols <- setNames(c("#E7B3BC", "#8FB3D9", "#B7D98C", "#C9B3D9",
                           "#E8794A", "#2A9D8F", "#B8B8B8", "#C0392B"), 1:params$n_clusters)
```

## Top clusters and genes

The top clusters are the `r params$n_top_clusters` clusters with the highest mean log2 fold change at
`r params$rank_medium` `r params$rank_time` h vs 2 h. In each of them, the top genes are those significantly
up-regulated in that comparison (padj < 0.05, log2FC ≥ 1), ranked by log2 fold change.

```{r top-genes}
de <- merge(de, data.table(gene = names(cl), cluster = cl), by = "gene")
# 2 h is the reference, so its log2FC vs 2 h is zero
de <- rbind(unique(de[, .(gene, medium, cluster)])[, `:=`(time = times[1], log2FC = 0, sig = FALSE)],
            de, fill = TRUE)

rank_de <- de[medium == params$rank_medium & time == params$rank_time]
cluster_rank <- rank_de[, .(mean_log2FC = mean(log2FC), n_up = sum(sig & log2FC > 0), n_genes = .N),
                        by = cluster][order(-mean_log2FC)]
cluster_rank
top_clusters <- head(cluster_rank$cluster, params$n_top_clusters)

top_genes <- rank_de[cluster %in% top_clusters & sig & log2FC > 0][
  order(-log2FC), head(.SD, params$n_top_genes), by = cluster]
gff <- setNames(sub("_A$", "", clusters$gff_id), clusters$gene)
top_genes[, label := ifelse(grepl("^CAALFM_", gene), gff[gene], gene)]
top_genes[, .(cluster, gene, label, log2FC = round(log2FC, 2), padj = signif(padj, 2))]
fwrite(top_genes, out_path("Figure_4C_top_genes.csv"))
```

## Panel B: heatmap

log2 fold change of each time point vs 2 h within each medium (2 h is 0 by definition). Rows follow
the dendrogram; the three top clusters are labelled in bold.

```{r panel-b, fig.width = 10, fig.height = 13}
lfc <- dcast(de[medium %in% media], gene ~ medium + time, value.var = "log2FC")
lfc_mat <- as.matrix(lfc[, -1]); rownames(lfc_mat) <- lfc$gene
lfc_mat <- lfc_mat[rownames(z), group_levels]
col_medium <- factor(sub("_.*", "", colnames(lfc_mat)), levels = media)
medium_cols <- c(Rich = "#D9934E", Minimal = "#C7B3DB")

# With a dendrogram and row_split = k, ComplexHeatmap cuts the tree into k slices in dendrogram order
slice_clusters <- unique(cl[hc$order])
row_titles <- paste0(slice_clusters, ". ", cluster_annotation[as.character(slice_clusters)])

ht <- Heatmap(
  lfc_mat, name = "log2FC",
  col = colorRamp2(c(-params$lfc_limit, 0, params$lfc_limit), c("#2C3E99", "white", "#E31A1C")),
  cluster_rows = as.dendrogram(hc), row_split = params$n_clusters, row_gap = unit(1.2, "mm"),
  row_dend_width = unit(22, "mm"), show_row_names = FALSE,
  row_title = row_titles, row_title_side = "right", row_title_rot = 0,
  row_title_gp = gpar(fontsize = 11,
                      fontface = ifelse(slice_clusters %in% top_clusters, "bold", "plain")),
  cluster_columns = FALSE, column_split = col_medium, column_gap = unit(1.5, "mm"),
  column_title_side = "bottom", column_title_gp = gpar(fontsize = 13, fontface = "bold"),
  column_labels = paste(sub(".*_", "", colnames(lfc_mat)), "h"), column_names_rot = 90,
  column_names_centered = TRUE, column_names_gp = gpar(fontsize = 11),
  left_annotation = rowAnnotation(
    Cluster = factor(cl[rownames(lfc_mat)], levels = 1:params$n_clusters),
    col = list(Cluster = cluster_cols), show_legend = FALSE, show_annotation_name = TRUE,
    annotation_name_side = "bottom", annotation_name_gp = gpar(fontsize = 11, fontface = "bold")),
  top_annotation = HeatmapAnnotation(
    Medium = col_medium, col = list(Medium = medium_cols), show_legend = FALSE,
    show_annotation_name = FALSE, simple_anno_size = unit(3, "mm")),
  heatmap_legend_param = list(title = expression(log[2]*FC~vs~2~h), direction = "vertical",
                              legend_height = unit(40, "mm"), title_gp = gpar(fontsize = 11, fontface = "bold")),
  border = TRUE, use_raster = TRUE, raster_quality = 4
)
ht_draw <- draw(ht, heatmap_legend_side = "left", padding = unit(c(2, 2, 2, 28), "mm"))
# The slice titles assume the slices follow the dendrogram order; check that they do
stopifnot(all(vapply(seq_along(slice_clusters), function(i)
  all(cl[rownames(lfc_mat)[row_order(ht_draw)[[i]]]] == slice_clusters[i]), logical(1))))
```

## Panel C: top genes over time

```{r panel-c, fig.width = 6, fig.height = 12}
cluster_title <- function(k, wrap = 60) {
  sprintf("<span style='color:%s'>&#9632;</span> **Cluster %s**<br>%s",
          cluster_cols[as.character(k)], k,
          gsub("\n", "<br>", stringr::str_wrap(cluster_annotation[as.character(k)], wrap)))
}
traj <- de[medium == params$rank_medium & gene %in% top_genes$gene]
traj <- merge(traj, top_genes[, .(gene, label, rank_lfc = log2FC)], by = "gene")
traj[, time_f := factor(time, levels = times)]
traj[, facet := factor(cluster_title(cluster), levels = cluster_title(top_clusters))]
traj[, label := factor(label, levels = unique(label[order(cluster, -rank_lfc)]))]

gene_shapes <- rep(c(16, 15, 17, 18, 4, 3, 8, 1, 0, 2), length.out = params$n_top_genes)
traj[, gene_idx := match(label, unique(label)), by = cluster]
traj[, gene_idx := (gene_idx - 1) %% params$n_top_genes + 1]
end_labels <- traj[time == max(times)]
grey_ramp <- grey.colors(params$n_top_genes, start = 0.05, end = 0.65)

p_c <- ggplot(traj, aes(time_f, log2FC, group = gene)) +
  annotate("rect", xmin = which(times == params$late_time) - 0.5, xmax = Inf, ymin = -Inf, ymax = Inf,
           fill = "grey92") +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_line(aes(colour = factor(gene_idx)), linewidth = 0.6) +
  geom_point(aes(colour = factor(gene_idx), shape = factor(gene_idx)), size = 1.8) +
  geom_text_repel(data = end_labels, aes(label = label, colour = factor(gene_idx)),
                  size = 3.2, fontface = "italic", direction = "y", hjust = 0,
                  nudge_x = 0.35, xlim = c(length(times) + 0.3, NA), segment.size = 0.2,
                  segment.colour = "grey70", min.segment.length = 0, box.padding = 0.08) +
  scale_colour_manual(values = grey_ramp, guide = "none") +
  scale_shape_manual(values = gene_shapes, guide = "none") +
  scale_x_discrete(expand = expansion(add = c(0.4, 2.6))) +
  facet_wrap(~ facet, ncol = 1, scales = "free_y") +
  coord_cartesian(clip = "off") +
  labs(x = "Time (h)", y = bquote(log[2]~fold~change~vs~"2 h ("*.(params$rank_medium)*")")) +
  theme_classic(base_size = 12) +
  theme(strip.background = element_blank(), strip.text = element_markdown(hjust = 0, size = 11, lineheight = 1.15),
        panel.spacing.y = unit(4, "mm"))
p_c
```

## Panel D: autophagy

Autophagy genes are spread over several clusters (`r paste(sprintf("%s: C%s", params$autophagy_genes, cl[params$autophagy_genes]), collapse = ", ")`),
so none of the top clusters in panel C is enriched for autophagy terms. They are enriched, however, among the genes
up-regulated at each time point vs 2 h, which is shown here with the ORA of those gene sets.
Time points without a significant term (BH-adjusted p ≥ 0.05) are shown as open grey circles.
The gene trajectories on the right are shown for rich medium only.

```{r panel-d, fig.width = 16, fig.height = 6}
ora_time[, c("medium", "time", "direction") :=
           tstrsplit(sub("^(\\w+) (\\d+)h vs 2h (up|down)$", "\\1|\\2|\\3", Cluster), "|", fixed = TRUE)]
ora_time[, time := as.integer(time)]
ora_time[, has_gene := grepl(sprintf("(^|/)%s(/|$)", params$highlight_gene), geneID)]
auto <- ora_time[direction == "up" & ID %in% params$autophagy_terms]
auto[, .(medium, time, Description, Count, FoldEnrichment = round(FoldEnrichment, 2),
         p.adjust = signif(p.adjust, 2), has_gene)][order(Description, medium, time)]
fwrite(auto[, .(medium, time, ID, Description, Count, GeneRatio, FoldEnrichment, p.adjust, has_gene)],
       out_path("Figure_4D_autophagy_ORA.csv"))

term_names <- setNames(unique(auto[, .(ID, Description)])$Description, unique(auto[, .(ID, Description)])$ID)
stopifnot(all(params$autophagy_terms %in% names(term_names)))
grid_d <- CJ(medium = media, time = times[-1], ID = params$autophagy_terms)
grid_d <- merge(grid_d, auto[, .(medium, time, ID, Count, p.adjust, has_gene)],
                by = c("medium", "time", "ID"), all.x = TRUE)
grid_d[, `:=`(medium = factor(medium, levels = media), time_f = factor(time, levels = times[-1]),
              term = factor(ID, levels = rev(params$autophagy_terms)))]
term_labels <- setNames(ifelse(params$autophagy_terms == params$autophagy_highlight,
                               sprintf("<b style='color:#B2182B'>%s</b>", term_names[params$autophagy_terms]),
                               term_names[params$autophagy_terms]), params$autophagy_terms)
late_x <- which(times[-1] == params$late_time)
hl_y <- which(levels(grid_d$term) == params$autophagy_highlight)

p_d1 <- ggplot(grid_d, aes(time_f, term)) +
  geom_blank() +  # trains the x scale on all time points, in order, before the subset layers
  annotate("rect", xmin = late_x - 0.5, xmax = Inf, ymin = -Inf, ymax = Inf, fill = "grey92") +
  annotate("rect", xmin = late_x - 0.5, xmax = Inf, ymin = hl_y - 0.45, ymax = hl_y + 0.45,
           fill = "#FBE3E3", colour = "#B2182B", linewidth = 0.4) +
  geom_point(data = grid_d[is.na(p.adjust)], shape = 1, size = 1.5, colour = "grey65") +
  geom_point(data = grid_d[!is.na(p.adjust)], aes(size = Count, fill = -log10(p.adjust)),
             shape = 21, colour = "grey20") +
  scale_y_discrete(labels = term_labels) +
  scale_fill_gradient(low = "#FDD9B5", high = "#8C2D04", name = expression(-log[10]~p[adj])) +
  scale_size_area(max_size = 7, name = "Genes") +
  facet_wrap(~ medium, nrow = 1) +
  labs(x = "Time (h), up-regulated vs 2 h", y = NULL) +
  theme_bw(base_size = 12) +
  theme(axis.text.y = element_markdown(size = 10.5), strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 12), panel.grid.minor = element_blank(),
        legend.position = "right")

traj_d <- de[medium == "Rich" & gene %in% params$autophagy_genes]  # minimal medium shown only in the ORA panel
stopifnot(setequal(unique(traj_d$gene), params$autophagy_genes))
traj_d[, `:=`(medium = factor(medium, levels = "Rich"), time_f = factor(time, levels = times),
              is_hl = gene == params$highlight_gene,
              filled = sig == TRUE | time == times[1])]  # 2 h is the reference, not a test
traj_d[, label := ifelse(is_hl, sprintf("%s (Atg8)", gene), gene)]
end_d <- traj_d[time == max(times)]

p_d2 <- ggplot(traj_d, aes(time_f, log2FC, group = gene)) +
  annotate("rect", xmin = which(times == params$late_time) - 0.5, xmax = Inf, ymin = -Inf, ymax = Inf,
           fill = "grey92") +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_line(data = traj_d[is_hl == FALSE], colour = "grey55", linewidth = 0.8) +
  geom_point(data = traj_d[is_hl == FALSE & filled == TRUE], colour = "grey55", size = 2.2) +
  geom_point(data = traj_d[is_hl == FALSE & filled != TRUE], colour = "grey55", size = 2.2, shape = 1, stroke = 0.8) +
  geom_line(data = traj_d[is_hl == TRUE], colour = "#B2182B", linewidth = 1.8) +
  geom_point(data = traj_d[is_hl == TRUE & filled == TRUE], colour = "#B2182B", size = 3.4) +
  geom_point(data = traj_d[is_hl == TRUE & filled != TRUE], colour = "#B2182B", size = 3.4, shape = 1, stroke = 1) +
  geom_text_repel(data = end_d, aes(label = label, colour = is_hl, fontface = ifelse(is_hl, "bold.italic", "italic")),
                  size = 4.2, direction = "y", hjust = 0, nudge_x = 0.35,
                  xlim = c(length(times) + 0.3, NA), segment.size = 0.2, segment.colour = "grey70",
                  min.segment.length = 0, box.padding = 0.08) +
  scale_colour_manual(values = c(`FALSE` = "grey35", `TRUE` = "#B2182B"), guide = "none") +
  scale_x_discrete(expand = expansion(add = c(0.4, 2.4))) +
  facet_wrap(~ medium, nrow = 1) +
  coord_cartesian(clip = "off") +
  labs(x = "Time (h)", y = expression(log[2]~fold~change~vs~"2 h"),
       caption = expression(Open~points*":"~p[adj] >= 0.05)) +
  theme_classic(base_size = 14) +
  theme(strip.background = element_blank(), strip.text = element_text(face = "bold", size = 14),
        plot.caption = element_text(size = 10.5, colour = "grey40"))

p_d1 + p_d2 + plot_layout(widths = c(1.3, 1))
```

## Assembled figure

```{r assemble, fig.width = 16, fig.height = 25}
panel_a <- wrap_elements(full = rasterGrob(readPNG(data_path("Figure_4A_schematic.png")),
                                           interpolate = TRUE))
panel_b <- wrap_elements(full = grid.grabExpr(
  draw(ht, heatmap_legend_side = "left", padding = unit(c(2, 2, 2, 28), "mm")),
  width = 10.5, height = 13))

row_bc <- (panel_b | p_c + theme(plot.margin = margin(5, 5, 5, 15))) +
  plot_layout(widths = c(10.5, 5.5))
row_d <- (p_d1 | p_d2) + plot_layout(widths = c(1.3, 1))
fig4 <- (panel_a / row_bc / wrap_elements(full = row_d)) +
  plot_layout(heights = c(5.2, 13, 5.6)) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 22, face = "bold"))
fig4

ggsave(out_path("Figure_4.pdf"), fig4, width = 16, height = 25, device = cairo_pdf)
ggsave(out_path("Figure_4.png"), fig4, width = 16, height = 25, dpi = 300)
```

## Session info

```{r session-info}
sessionInfo()
```
