---
title: "Supplementary Figure 9: GO over-representation analysis of temporal expression clusters"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
  n_clusters: 8
  show_category: 5         # terms per cluster, as dotplot(showCategory = 5) in the analysis
  label_width: 50          # wrap GO term names, as dotplot(label_format = 50) in the analysis
  late_time: 24            # first quiescent time point (shaded)
  p_cap: 1.0e-20           # colour scale floor for adjusted p (smaller values shown at the cap)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 13, fig.height = 15)
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(ggtext)
  library(patchwork)
})
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(f) file.path(params$data_dir, f)
out_path <- function(...) file.path(params$out_dir, ...)
```

## Figure legend

**Supplementary Figure 9. GO over-representation analysis of the temporal expression clusters.**
Time-responsive genes (likelihood-ratio test padj < 0.05 in rich or minimal medium) were grouped into
`r params$n_clusters` clusters by hierarchical clustering of their z-scored mean expression in each medium
and time point (Figure 4B). Top, expression pattern of each cluster in rich and minimal medium: median z-score
(line) and interquartile range (band); n, genes per cluster. Time is on a log~2~ scale and the shaded area
marks the quiescent phase (`r params$late_time`–72 h). Bottom, GO biological process terms over-represented in
each cluster (hypergeometric test against the expressed, annotated genes; Benjamini–Hochberg adjusted
p < 0.05). The `r params$show_category` most significant terms of each cluster are shown, together with
their enrichment in any other cluster. Dot size is the fraction of the cluster's annotated genes in the
term (gene ratio) and colour the adjusted p-value (values below 10^-20^ are shown at 10^-20^). The
shaded block in each column marks the terms selected for that cluster.

## Input data

The inputs in `data/` are outputs of
`CandidaTFKO5_WT_72h/analysis/WT_Rich_vs_Minimal_timecourse_DESeq2_ORA.Rmd`
(`new_72h_mode = "add"`, `n_clusters = 8`, GO BP). Figure 4 uses the same files. This figure combines the
"Expression patterns over time" and "ORA of temporal clusters" plots of that analysis.

| File | Description |
|---|---|
| `vst_batch_corrected.csv.gz` | DESeq2 VST, sequencing batch removed, 46 libraries (WT, rich and minimal, 2–72 h) |
| `timecourse_gene_clusters.csv` | Cluster assignment of the time-responsive genes |
| `ORA_GO_timecourse_clusters.csv` | GO BP over-representation per cluster (`compareCluster(fun = enricher)`, BH-adjusted p < 0.05) |

```{r load}
vst <- fread(data_path("vst_batch_corrected.csv.gz"), data.table = FALSE)
vmat <- as.matrix(vst[, -1]); rownames(vmat) <- vst[[1]]
clusters <- fread(data_path("timecourse_gene_clusters.csv"), data.table = FALSE)
ora <- fread(data_path("ORA_GO_timecourse_clusters.csv"))

# Sample groups. CandidaTFKO5_WT_72h-* are the Rich 72 h libraries from run HJT3JDRX7.
sample_medium <- ifelse(grepl("^WT_C_", colnames(vmat)), "Minimal", "Rich")
sample_time <- ifelse(grepl("^CandidaTFKO5_WT_72h", colnames(vmat)), 72L,
                      as.integer(sub("^WT_[YC]_(\\d+)-\\d$", "\\1", colnames(vmat))))
times <- sort(unique(sample_time))
media <- c("Rich", "Minimal")
group_levels <- paste(rep(media, each = length(times)), times, sep = "_")
sample_group <- factor(paste(sample_medium, sample_time, sep = "_"), levels = group_levels)
table(sample_medium, sample_time)
```

## Cluster expression patterns

As in the analysis, the VST values of the time-responsive genes are averaged per medium × time group and
z-scored across all 14 groups, so rich and minimal are on the same scale. The clustering is rebuilt and
checked against the saved assignment.

```{r clusters}
gmeans <- sapply(group_levels, function(g) rowMeans(vmat[clusters$gene, sample_group == g, drop = FALSE]))
z <- t(scale(t(gmeans)))
z <- z[complete.cases(z), ]
hc <- hclust(dist(z), method = "ward.D2")
cl <- cutree(hc, k = params$n_clusters)
peak <- tapply(seq_along(cl), cl, function(i) which.max(colMeans(z[i, grep("^Rich", colnames(z)), drop = FALSE])))
cl <- match(cl, order(peak)); names(cl) <- rownames(z)
stopifnot(identical(unname(cl[clusters$gene]), clusters$cluster))
n_genes <- table(factor(cl, levels = 1:params$n_clusters))
n_genes

zl <- melt(data.table(gene = rownames(z), z), id.vars = "gene", variable.name = "group",
           value.name = "z", variable.factor = FALSE)
zl[, c("medium", "time") := tstrsplit(group, "_", fixed = TRUE)]
zl[, `:=`(time = as.integer(time), cluster = cl[gene])]
traj <- zl[, .(med = median(z), lo = quantile(z, 0.25), hi = quantile(z, 0.75)), by = .(cluster, medium, time)]
fwrite(traj[order(cluster, medium, time)], out_path("Supplementary_Figure_09_cluster_patterns.csv"))
```

## GO term selection

Terms are selected as in `enrichplot::dotplot()` for a `compareCluster` result with
showCategory = `r params$show_category`: all results are sorted by p-value, the top
`r params$show_category` of each cluster are kept, and every significant result for those terms, in any
cluster, is shown. Each term is listed under the first cluster that selected it, ordered by adjusted
p-value, so the terms of each cluster form a block.

```{r terms}
ora[, cluster := as.integer(sub("^C", "", Cluster))]
ora[, gene_ratio := vapply(strsplit(GeneRatio, "/", fixed = TRUE),
                           function(x) as.numeric(x[1]) / as.numeric(x[2]), numeric(1))]
top <- ora[order(pvalue)][, head(.SD, params$show_category), by = cluster]
term_order <- top[order(cluster, p.adjust), .(ID, home = cluster)][!duplicated(ID)]
dots <- merge(ora[ID %in% term_order$ID], term_order, by = "ID")
dots[, label := stringr::str_wrap(Description, params$label_width)]
dots[, term := factor(ID, levels = rev(term_order$ID))]
term_labels <- setNames(unique(dots[, .(ID, label)])$label, unique(dots[, .(ID, label)])$ID)
cat(nrow(term_order), "terms;", nrow(dots), "cluster-term results shown\n")
dots[order(home, term, cluster), .(home, cluster, Description, GeneRatio, Count,
                                   FoldEnrichment = round(FoldEnrichment, 1), p.adjust = signif(p.adjust, 2))]
fwrite(dots[order(home, -as.integer(term), cluster),
            .(term_cluster = home, cluster, ID, Description, GeneRatio, BgRatio, Count, FoldEnrichment,
              pvalue, p.adjust, geneID)],
       out_path("Supplementary_Figure_09_ORA_terms_shown.csv"))
```

## Figure

The cluster columns of the two panels are aligned. Cluster colours follow Figure 4B.

```{r figure, fig.width = 13, fig.height = 15}
medium_cols <- c(Rich = "#D9934E", Minimal = "#8C6BB1")  # as Supplementary Figure 8
cluster_cols <- setNames(c("#E7B3BC", "#8FB3D9", "#B7D98C", "#C9B3D9",
                           "#E8794A", "#2A9D8F", "#B8B8B8", "#C0392B"), 1:params$n_clusters)  # Figure 4
cluster_strip <- setNames(sprintf("<span style='color:%s'>&#9632;</span> **Cluster %d**<br>n = %s",
                                  cluster_cols, 1:params$n_clusters, format(as.integer(n_genes), big.mark = ",")),
                          1:params$n_clusters)
facet_theme <- theme(strip.background = element_blank(), strip.text = element_markdown(size = 11, lineheight = 1.15),
                     panel.spacing.x = unit(2.5, "mm"))

traj[, `:=`(medium = factor(medium, levels = media), facet = factor(cluster, levels = 1:params$n_clusters))]
p_top <- ggplot(traj, aes(time, med, colour = medium, fill = medium)) +
  annotate("rect", xmin = sqrt(params$late_time * max(times[times < params$late_time])), xmax = Inf,
           ymin = -Inf, ymax = Inf, fill = "grey92") +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.25, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = medium_cols, name = "Medium") +
  scale_fill_manual(values = medium_cols, name = "Medium") +
  scale_x_continuous(trans = "log2", breaks = c(2, 8, 24, 72)) +
  facet_wrap(~ facet, nrow = 1, labeller = as_labeller(cluster_strip)) +
  labs(x = "Time (h)", y = "Expression (z-score)") +
  theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank(), axis.text.x = element_text(size = 8.5)) +
  facet_theme

dots[, facet := factor(cluster, levels = 1:params$n_clusters)]
# Pale band behind each cluster's own block of terms, in its own column
blocks <- dots[cluster == home, .(ymin = min(as.integer(term)) - 0.5, ymax = max(as.integer(term)) + 0.5),
               by = .(home)][, facet := factor(home, levels = 1:params$n_clusters)]
p_cap <- params$p_cap
p_bottom <- ggplot(dots, aes(x = 1, y = term)) +
  geom_blank() +  # sets the discrete y scale before the continuous block rectangles
  geom_rect(data = blocks, aes(xmin = -Inf, xmax = Inf, ymin = ymin, ymax = ymax, fill = NULL),
            inherit.aes = FALSE, fill = alpha(cluster_cols[as.character(blocks$home)], 0.25)) +
  geom_point(aes(size = gene_ratio, fill = pmax(p.adjust, p_cap)), shape = 21, colour = "grey25", stroke = 0.3) +
  scale_y_discrete(labels = term_labels) +
  scale_x_continuous(breaks = NULL, expand = expansion(add = 1)) +
  scale_fill_gradientn(colours = c("#7F0000", "#D7301F", "#FC8D59", "#FDD49E", "#FFF7EC"),
                       trans = "log10", limits = c(p_cap, 0.05), name = "Adjusted p",
                       breaks = 10^-c(20, 15, 10, 5, 2), labels = c(sprintf("\u2264 1e-%d", 20), "1e-15", "1e-10", "1e-5", "0.01"),
                       guide = guide_colourbar(reverse = TRUE, barheight = unit(40, "mm"))) +
  scale_size(range = c(1.8, 7.5), name = "Gene ratio") +
  facet_wrap(~ facet, nrow = 1, drop = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 12) +
  theme(axis.text.y = element_text(size = 10, lineheight = 0.9), panel.grid.major.y = element_line(colour = "grey92")) +
  facet_theme + theme(strip.text = element_blank())

n_terms <- nlevels(dots$term)
fig <- (p_top / p_bottom) +
  plot_layout(heights = c(2.2, n_terms * 0.27), guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 18, face = "bold"))
fig_height <- 3.4 + n_terms * 0.27
fig

ggsave(out_path("Supplementary_Figure_09.pdf"), fig, width = 13, height = fig_height, device = cairo_pdf)
ggsave(out_path("Supplementary_Figure_09.png"), fig, width = 13, height = fig_height, dpi = 300)
```

## Session info

```{r session-info}
sessionInfo()
```
