---
title: "Supplementary Figures 15-20: PI/SYTO9 gating of SC5314 exposed to caspofungin, micafungin and amphotericin B"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
  experiment: "Casp_Mica"   # "Casp_Mica" (Supp. Figs 15, 16, 18, 19) or "AmpB" (Supp. Figs 17, 20)
  fcs_dir: "FCS_files"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(f) file.path(params$data_dir, params$experiment, f)
out_path <- function(...) file.path(params$out_dir, ...)

settings <- list(
  Casp_Mica = list(version_name = "OzanImir_20250911_v1",
                   details = "Vignette-Experiment-Details.csv",
                   markers = "Vignette-Experiment-Markers.csv",
                   rows_per_page = 6L, reverse_order = FALSE),
  AmpB = list(version_name = "OzanImir_20251009_v2",
              details = "161025-Experiment-Details.csv",
              markers = "161025-Experiment-Markers.csv",
              rows_per_page = 7L, reverse_order = TRUE)
)[[params$experiment]]
```

## Figure legend

**Supplementary Figure 15.** Cell viability quantification to validate accuracy of PI/SYTO9 staining
quantification with caspofungin exposed proliferative cells. A) untreated; B–F) 0.001, 0.01, 0.1, 1
and 10 ug/mL caspofungin. Each panel shows PI and SYTO9 distribution functions as well as the gated
PI/SYTO9 2D quantification; cells grown in rich media.

**Supplementary Figure 16.** As Supplementary Figure 15, with micafungin exposed proliferative cells.

**Supplementary Figure 17.** As Supplementary Figure 15, with amphotericin B exposed proliferative
cells (A untreated; B–G 0.001, 0.01, 0.1, 1, 10 and 20 ug/mL amphotericin B).

**Supplementary Figure 18.** As Supplementary Figure 15, with caspofungin exposed quiescent cells.

**Supplementary Figure 19.** As Supplementary Figure 15, with micafungin exposed quiescent cells.

**Supplementary Figure 20.** As Supplementary Figure 17, with amphotericin B exposed quiescent cells.

(Full legend text is in the folder README.)

## Input data

Raw FCS files are **not distributed**, so this whole analysis is documented but not evaluated
(`eval = FALSE`); rendering only prints the experiment details. Set `params$experiment` to choose
the data set; files are read from `data/<experiment>/`.

| File | Description |
|---|---|
| `data/Casp_Mica/Vignette-Experiment-Details.csv` | 96 FCS files: Strain1 (SC5314) and Strain2, 5FC/AmpB/caspofungin/micafungin, 0–10 ug/mL, exponential and quiescent |
| `data/Casp_Mica/sample_sheet_vignette_simple.csv` | Sample sheet (name, strain, drug, dose, growth) merged into the experiment details |
| `data/Casp_Mica/Vignette-Experiment-Markers.csv` | Channel/marker table (B12-A = SYTO9, YG4-A = PI) |
| `data/Casp_Mica/cytek_gating_OzanImir_20250911_v1.csv` | Gating template (Cells, Single_cells, Live_cells, Dead_cells) |
| `data/AmpB/161025-Experiment-Details.csv` | 14 FCS files: SC5314 untreated and 0.001–20 ug/mL amphotericin B, exponential and quiescent (file names only, no metadata columns) |
| `data/AmpB/161025-Experiment-Markers.csv` | Channel/marker table (B7-A, B12-A = SYTO9, YG4-A = PI) |
| `data/AmpB/cytek_gating_OzanImir_20251009_v2.csv` | Gating template (Cells, Single_cells, Live_cells, Dead_cells) |

```{r experiment-details}
settings[c("version_name", "rows_per_page", "reverse_order")]
read.csv(data_path(settings$details))
```

## Gating from FCS (not evaluated; FCS not distributed)

```{r gating-libraries, eval = FALSE}
library(cowplot)
library(flowCore)
library(flowWorkspace)
library(openCyto)
library(CytoExploreR)
library(tidyverse)
```

Load the FCS files (expected in `params$fcs_dir`) as a gating set with the shipped experiment
details and markers:

```{r gating-load, eval = FALSE}
my_gating_set <- cyto_setup(path = params$fcs_dir, restrict = TRUE, select = "fcs",
                            details = data_path(settings$details),
                            markers = data_path(settings$markers))
cyto_details(my_gating_set)
```

Arcsinh transforms (cofactor 150 for SYTO9 and PI, 500 for FSC/SSC):

```{r gating-transform, eval = FALSE}
SYTO9_trans <- cyto_transformer_arcsinh(my_gating_set, channels = c("B12-A"), cofactor = 150)
PI_trans <- cyto_transformer_arcsinh(my_gating_set, channels = c("YG4-A"), cofactor = 150)
FSC_SSC_trans <- cyto_transformer_arcsinh(my_gating_set,
                                          channels = c("FSC-A", "FSC-H", "SSC-A", "SSC-H"),
                                          cofactor = 500)
combined_trans <- cyto_transformer_combine(SYTO9_trans, PI_trans, FSC_SSC_trans)
transformed_gating_set <- cyto_transform(my_gating_set, trans = combined_trans)
```

Apply the gating template (Cells on FSC-A/SSC-A, Single_cells on SSC-A/SSC-H, Live_cells and
Dead_cells on YG4-A/B12-A; gates were originally drawn with `cyto_gate_draw()`):

```{r gating-apply, eval = FALSE}
cyto_gatingTemplate_apply(transformed_gating_set,
                          gatingTemplate = data_path(paste0("cytek_gating_", settings$version_name, ".csv")))
```

PI and SYTO9 histograms, one row per sample (6 rows per page for Casp_Mica; 7 rows per page in
reversed sample order for AmpB):

```{r gating-histograms, eval = FALSE}
theme_set(theme_cowplot())
gs <- transformed_gating_set
rows_per_page <- settings$rows_per_page

get_df <- function(gs, i, chans = c("YG4-A", "B12-A")) {
  cf <- gh_pop_get_data(gs[[i]], "root", inverse.transform = TRUE)
  ff <- cytoframe_to_flowFrame(cf)
  m <- exprs(ff)
  df <- as.data.frame(m[, chans, drop = FALSE], check.names = FALSE)
  names(df) <- chans
  df$PI <- df[[chans[1]]]
  df$SYTO9 <- df[[chans[2]]]
  df
}

# log10 histogram when positive values exist, otherwise linear
plot_hist_log_if_positive <- function(df, var, xlabel) {
  x_max <- 1e6
  if (any(df[[var]] > 0, na.rm = TRUE)) {
    df_plot <- df[df[[var]] > 0 & is.finite(df[[var]]), , drop = FALSE]
    ggplot(df_plot, aes_string(var)) +
      geom_histogram(bins = 300, fill = "steelblue", color = "black", alpha = 0.6) +
      scale_x_log10(limits = c(1, x_max)) +
      ggplot2::annotation_logticks(sides = "b") +
      coord_cartesian(xlim = c(1, x_max)) +
      labs(x = xlabel, y = "Count")
  } else {
    ggplot(df, aes_string(var)) +
      geom_histogram(bins = 300, fill = "steelblue", color = "black", alpha = 0.6) +
      coord_cartesian(xlim = c(0, x_max)) +
      labs(x = xlabel, y = "Count")
  }
}

n_samples <- length(gs)
n_pages <- ceiling(n_samples / rows_per_page)
sample_order <- if (settings$reverse_order) rev(seq_len(n_samples)) else seq_len(n_samples)

for (page in seq_len(n_pages)) {
  idx <- sample_order[((page - 1) * rows_per_page + 1):min(page * rows_per_page, n_samples)]
  row_plots <- vector("list", length(idx))
  for (k in seq_along(idx)) {
    i <- idx[k]
    df <- get_df(gs, i, c("YG4-A", "B12-A"))
    p_pi <- plot_hist_log_if_positive(df, "PI", "PI")
    p_sy <- plot_hist_log_if_positive(df, "SYTO9", "SYTO9")
    row_core <- plot_grid(p_pi, p_sy, ncol = 2, align = "hv", axis = "tblr")
    row_plots[[k]] <- plot_grid(row_core, labels = LETTERS[k], label_size = 18,
                                label_fontface = "bold", label_x = 0, hjust = 0)
  }
  page_grid <- plot_grid(plotlist = row_plots, ncol = 1)
  ggsave(filename = out_path(sprintf("PI_SYTO9_grid_2x%d_p%02d.pdf", rows_per_page, page)),
         plot = page_grid, device = grDevices::pdf,
         width = 10.5, height = 18, units = "in", useDingbats = FALSE)
}
```

Gated PI x SYTO9 scatter plots with Live_cells / Dead_cells percentages (right panels):

```{r gating-scatter, eval = FALSE}
save_gating_png <- function(gs, samp, outfile) {
  grDevices::png(filename = outfile, width = 6, height = 4, units = "in", res = 300)
  on.exit(grDevices::dev.off(), add = TRUE)
  cyto_plot(gs, parent = "Single_cells", channels = c("YG4-A", "B12-A"),
            axes_limits = "data", xlab = "PI", ylab = "SYTO9", alias = "")
}

for (i in seq_along(transformed_gating_set)) {
  samp <- flowWorkspace::sampleNames(transformed_gating_set)[i]
  outfile <- out_path(sprintf("PI_SYTO9_gating_%s.png", samp))
  try(save_gating_png(transformed_gating_set[i], samp, outfile), silent = TRUE)
}
```

## Session info

```{r session-info}
sessionInfo()
```
