---
title: "Supplementary Figure 14: PI/SYTO9 viability controls with heat-killed cells"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
  fcs_dir: "FCS_files"
  version_name: "OzanImir_20250918_v2"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 5)
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(ggplot2)
})
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(f) file.path(params$data_dir, f)
out_path <- function(...) file.path(params$out_dir, ...)
```

## Figure legend

**Supplementary Figure 14. Validation of PI/SYTO9 staining and flow cytometry to measure viability
of proliferative and quiescent cells.** In each panel the distribution of PI (left panel), SYTO9
(middle panel) and the scatter plot of PI versus SYTO9 used to define flow cytometry gates is
displayed for A) untreated proliferative cells grown (100% viability expected), B) a 50:50 mixture of
untreated cells and cells exposed to 68℃ for 1 hour (50% viability expected) C) cells exposed to 68℃
for 1 hour (100% inviability expected) D) untreated quiescent cells (100% viability expected), E) a
50:50 mixture of untreated quiescent cells and quiescent cells exposed to 68℃ for 1 hour (50%
viability expected), and F) quiescent cells exposed to 68℃ for 1 hour (100% inviability expected).
For the experiment, all cells were grown in rich media.

## Input data

Raw FCS files are **not distributed**. The gating section documents how the figure panels were made
from FCS with CytoExploreR and is not evaluated. The summary section runs from values in this file.

| File | Description |
|---|---|
| `data/Vignette-Experiment-Details.csv` | CytoExploreR experiment details: FCS file name, expected viability (100/50/0), growth phase |
| `data/sample_sheet_vignette_simple.csv` | Sample sheet merged into the experiment details |
| `data/Vignette-Experiment-Markers.csv` | Channel/marker table (B12-A = SYTO9, YG4-A = PI) |
| `data/cytek_gating_OzanImir_20250918_v2.csv` | Gating template used (`version_name`); contains the Live_cells and Dead_cells gates only |
| `data/cytek_gating_OzanImir_20250918_v1.csv` | Earlier template; contains the Cells and Single_cells gates that v2's gates use as parent |

```{r experiment-details}
read.csv(data_path("Vignette-Experiment-Details.csv"))
```

## Gating from FCS (not evaluated; FCS not distributed)

Panels A–F correspond to the six FCS files in the order of `Vignette-Experiment-Details.csv`
(Exponential 100/50/0 % alive, then Quiescent 100/50/0 % alive). Channels: PI = YG4-A,
SYTO9 = B12-A.

```{r gating-libraries, eval = FALSE}
library(cowplot)
library(flowCore)
library(flowWorkspace)
library(openCyto)
library(CytoExploreR)
library(tidyverse)
```

Load the FCS files as a gating set (expects the FCS files in `params$fcs_dir`) and attach the
experiment details. The merge of the sample sheet into the experiment details was done once.

```{r gating-load, eval = FALSE}
my_gating_set <- cyto_setup(path = params$fcs_dir, restrict = TRUE, select = "fcs",
                            details = data_path("Vignette-Experiment-Details.csv"),
                            markers = data_path("Vignette-Experiment-Markers.csv"))

# One-time merge of the auto-generated experiment details with the sample sheet
# exp_details <- read_csv(data_path("Vignette-Experiment-Details.csv"))
# sample_sheet <- read_csv(data_path("sample_sheet_vignette_simple.csv"))
# left_join(exp_details, sample_sheet) %>% write_csv(data_path("Vignette-Experiment-Details.csv"))

cyto_details(my_gating_set)
```

Arcsinh transforms (cofactor 150 for SYTO9 and PI, 500 for FSC/SSC):

```{r gating-transform, eval = FALSE}
SYTO9_trans <- cyto_transformer_arcsinh(my_gating_set, channels = c("B12-A"), cofactor = 150)
PI_trans <- cyto_transformer_arcsinh(my_gating_set, channels = c("YG4-A"), cofactor = 150)
FSC_SSC_trans <- cyto_transformer_arcsinh(my_gating_set,
                                          channels = c("FSC-A", "FSC-H", "SSC-A", "SSC-H"),
                                          cofactor = 500)
combined_trans <- cyto_transformer_combine(SYTO9_trans, PI_trans, FSC_SSC_trans)
transformed_gating_set <- cyto_transform(my_gating_set, trans = combined_trans)
```

Apply the gating template (gates were originally drawn interactively with `cyto_gate_draw()`:
Cells on FSC-A/SSC-A, Single_cells on SSC-A/SSC-H, then Live_cells and Dead_cells on YG4-A/B12-A).
Note: the v2 template holds only Live_cells and Dead_cells (parent `Single_cells`); the Cells and
Single_cells gates are in the v1 template.

```{r gating-apply, eval = FALSE}
cyto_gatingTemplate_apply(transformed_gating_set,
                          gatingTemplate = data_path(paste0("cytek_gating_", params$version_name, ".csv")))
```

PI and SYTO9 histograms (left and middle panels), one row per sample, labelled A–F:

```{r gating-histograms, eval = FALSE}
theme_set(theme_cowplot())
gs <- transformed_gating_set
rows_per_page <- 6L

get_df <- function(gs, i, chans = c("YG4-A", "B12-A")) {
  cf <- gh_pop_get_data(gs[[i]], "root", inverse.transform = TRUE)
  ff <- cytoframe_to_flowFrame(cf)
  m <- exprs(ff)
  df <- as.data.frame(m[, chans, drop = FALSE], check.names = FALSE)
  names(df) <- chans
  df$PI <- df[[chans[1]]]
  df$SYTO9 <- df[[chans[2]]]
  df
}

# log10 histogram when positive values exist, otherwise linear
plot_hist_log_if_positive <- function(df, var, xlabel) {
  x_max <- 1e6
  if (any(df[[var]] > 0, na.rm = TRUE)) {
    df_plot <- df[df[[var]] > 0 & is.finite(df[[var]]), , drop = FALSE]
    ggplot(df_plot, aes_string(var)) +
      geom_histogram(bins = 300, fill = "steelblue", color = "black", alpha = 0.6) +
      scale_x_log10(limits = c(1, x_max)) +
      ggplot2::annotation_logticks(sides = "b") +
      coord_cartesian(xlim = c(1, x_max)) +
      labs(x = xlabel, y = "Count")
  } else {
    ggplot(df, aes_string(var)) +
      geom_histogram(bins = 300, fill = "steelblue", color = "black", alpha = 0.6) +
      coord_cartesian(xlim = c(0, x_max)) +
      labs(x = xlabel, y = "Count")
  }
}

n_samples <- length(gs)
n_pages <- ceiling(n_samples / rows_per_page)

for (page in seq_len(n_pages)) {
  idx <- ((page - 1) * rows_per_page + 1):min(page * rows_per_page, n_samples)
  row_plots <- vector("list", length(idx))
  for (k in seq_along(idx)) {
    i <- idx[k]
    df <- get_df(gs, i, c("YG4-A", "B12-A"))
    p_pi <- plot_hist_log_if_positive(df, "PI", "PI")
    p_sy <- plot_hist_log_if_positive(df, "SYTO9", "SYTO9")
    row_core <- plot_grid(p_pi, p_sy, ncol = 2, align = "hv", axis = "tblr")
    row_plots[[k]] <- plot_grid(row_core, labels = LETTERS[k], label_size = 18,
                                label_fontface = "bold", label_x = 0, hjust = 0)
  }
  grid_2x6 <- plot_grid(plotlist = row_plots, ncol = 1)
  ggsave(filename = out_path(sprintf("PI_SYTO9_grid_2x6_p%02d.pdf", page)),
         plot = grid_2x6, device = grDevices::pdf,
         width = 10.5, height = 18, units = "in", useDingbats = FALSE)
}
```

Gated PI x SYTO9 scatter plots with Live_cells / Dead_cells percentages (right panels):

```{r gating-scatter, eval = FALSE}
save_gating_png <- function(gs, samp, outfile) {
  grDevices::png(filename = outfile, width = 6, height = 4, units = "in", res = 300)
  on.exit(grDevices::dev.off(), add = TRUE)
  cyto_plot(gs, parent = "Single_cells", channels = c("YG4-A", "B12-A"),
            axes_limits = "data", xlab = "PI", ylab = "SYTO9", alias = "")
}

for (i in seq_along(transformed_gating_set)) {
  samp <- flowWorkspace::sampleNames(transformed_gating_set)[i]
  outfile <- out_path(sprintf("PI_SYTO9_gating_%s.png", samp))
  try(save_gating_png(transformed_gating_set[i], samp, outfile), silent = TRUE)
}
```

## Expected vs measured viability

Measured % live values were typed in from the gating of the heat-killing experiment
(original script `HeatKilling_Quantification.R`).

```{r quantification-data}
data <- tribble(
  ~Condition,       ~Group,        ~Expected_Live, ~Measured_Live,
  "Proliferative",  "100/0",       100,            95.40,
  "Proliferative",  "0/100",       0,              3.91,
  "Proliferative",  "50/50",       50,             44.44,
  "Quiescent",      "100/0",       100,            97.76,
  "Quiescent",      "0/100",       0,              1.70,
  "Quiescent",      "50/50",       50,             50.47
) %>%
  mutate(
    Expected_Dead = 100 - Expected_Live,
    Measured_Dead = 100 - Measured_Live
  )
data
```

```{r plot-live, fig.width = 6, fig.height = 5}
plot_live <- ggplot(data, aes(x = Expected_Live, y = Measured_Live, color = Condition)) +
  geom_point(size = 4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Proliferative" = "#F2766F", "Quiescent" = "#24BCBE"))
plot_live

ggsave(out_path("percent_viability_live.png"), plot_live, width = 6, height = 5, dpi = 300)
ggsave(out_path("percent_viability_live.pdf"), plot_live, width = 6, height = 5)
```

```{r plot-dead, fig.width = 6, fig.height = 5}
plot_dead <- ggplot(data, aes(x = Expected_Dead, y = Measured_Dead, color = Condition)) +
  geom_point(size = 4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Proliferative" = "#F2766F", "Quiescent" = "#24BCBE"))
plot_dead

ggsave(out_path("percent_viability_dead.png"), plot_dead, width = 6, height = 5, dpi = 300)
ggsave(out_path("percent_viability_dead.pdf"), plot_dead, width = 6, height = 5)
```

## Session info

```{r session-info}
sessionInfo()
```
