---
title: "Figure 1: Final yield versus initial glucose concentration in rich media"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(readxl)
  library(ggplot2)
})
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(f) file.path(params$data_dir, f)
out_path <- function(...) file.path(params$out_dir, ...)
```

## Figure legend

**Figure 1C (legend lettering; panel A in the current figure).** Final population yield (OD at
600nm) versus initial glucose concentration for rich media. The line indicates a linear model
fit to the data. The dashed circle indicates the concentration of glucose used in rich media for
all subsequent experiments.

## Input data

| File | Description |
|---|---|
| `Resubmission_GlucoseConcOD.xlsx` | Plate-reader endpoint OD600 export of SC5314 grown in rich media with 4%, 2%, 1% and 0.5% glucose (plate rows A–D, columns 9–12 = 4 biological replicates); row E = cell-free medium blank; rows F–H not part of this design |

Corrections: the mean of the four blank wells (row E) is subtracted from every reading (negative
values clipped to zero) and the result is multiplied by the dilution factor of the sample before
reading (`c(5, 10, 10, 10)` for 4%, 2%, 1%, 0.5% glucose). Glucose % (w/v) is converted to molar
as M = % × 10 / 180.16.

## Parse plate

```{r parse}
raw <- as.data.frame(read_excel(data_path("Resubmission_GlucoseConcOD.xlsx"),
                                sheet = 1, col_names = FALSE),
                     stringsAsFactors = FALSE)

hdr <- which(trimws(raw[[1]]) == "<>")
stopifnot("could not locate the '<>' plate header row" = length(hdr) == 1)

block      <- raw[(hdr + 1):(hdr + 8), ]           # plate rows A-H
row_labels <- trimws(block[[1]])
vals       <- block[, -1, drop = FALSE]
vals       <- vals[, colSums(!is.na(vals)) > 0, drop = FALSE]  # measured cols
plate      <- as.matrix(data.frame(lapply(vals, as.numeric),
                                 check.names = FALSE))
rownames(plate) <- row_labels
colnames(plate) <- paste0("Rep", seq_len(ncol(plate)))

knitr::kable(plate, caption = "Raw OD600 readings (plate rows A-H, 4 replicates)")
```

## Blank and dilution correction

```{r correct}
row_info <- data.frame(
  row       = c("A", "B", "C", "D", "E"),
  glucose   = c(4, 2, 1, 0.5, NA),            # % glucose; NA = no cells
  dilution  = c(5, 10, 10, 10, 1),            # dilution factor before reading
  condition = c("4% glucose", "2% glucose", "1% glucose",
                "0.5% glucose", "Cell-free control"),
  role      = c(rep("sample", 4), "blank"),
  stringsAsFactors = FALSE)

blank_mu <- mean(plate["E", ])                 # cell-free control mean

samples <- row_info[row_info$role == "sample", ]
long_df <- do.call(rbind, lapply(seq_len(nrow(samples)), function(i) {
  r <- samples$row[i]
  data.frame(
    glucose      = samples$glucose[i],
    dilution     = samples$dilution[i],
    replicate    = colnames(plate),
    OD_raw       = as.numeric(plate[r, ]),
    OD_corrected = pmax(as.numeric(plate[r, ]) - blank_mu, 0) *
                     samples$dilution[i],
    stringsAsFactors = FALSE)
}))

cat(sprintf("Blank (cell-free control, row E) mean OD600 = %.4f\n", blank_mu))

sum_df <- do.call(rbind, lapply(split(long_df, long_df$glucose), function(d) {
  data.frame(
    `Glucose (%)`       = d$glucose[1],
    `Dilution factor`   = d$dilution[1],
    `Mean raw OD`       = mean(d$OD_raw),
    `Mean corrected OD` = mean(d$OD_corrected),
    `SD corrected OD`   = sd(d$OD_corrected),
    n                   = nrow(d),
    check.names         = FALSE)
}))
sum_df <- sum_df[order(-sum_df$`Glucose (%)`), ]
knitr::kable(sum_df, row.names = FALSE, digits = 4,
             caption = "Blank- and dilution-corrected OD600 (n = 4)")
```

## Linear model

```{r fit}
MW_GLUCOSE        <- 180.16
long_df$glucose_M <- long_df$glucose * 10 / MW_GLUCOSE

fit <- lm(OD_corrected ~ glucose_M, data = long_df)
b0  <- coef(fit)[1]   # intercept
b1  <- coef(fit)[2]   # slope (OD600 per M glucose)
r2  <- summary(fit)$r.squared
p_val <- summary(fit)$coefficients[2, 4]

cat(sprintf("Slope:     %.4f OD600 per M glucose\n", b1))
cat(sprintf("R-squared: %.4f\n", r2))
cat(sprintf("p-value:   %.3e\n", p_val))
```

## Plot

```{r plot, fig.width=7, fig.height=5}
pt_color <- "#E69F00"

# fixed x-offsets so the 4 replicates are distinguishable
long_df$x_offset <- (as.integer(factor(long_df$replicate)) - 2.5) * 0.0003

p_glc <- ggplot(long_df, aes(glucose_M + x_offset, OD_corrected)) +
  geom_segment(aes(x = min(glucose_M), xend = max(glucose_M),
                   y = b0 + b1 * min(glucose_M),
                   yend = b0 + b1 * max(glucose_M)),
               linewidth = 1.0, colour = "black",
               inherit.aes = FALSE) +
  geom_point(size = 2.5, colour = pt_color, shape = 16) +
  scale_x_continuous(
    breaks = c(0.5, 1, 2, 4) * 10 / MW_GLUCOSE,
    labels = scales::number_format(accuracy = 0.001)) +
  coord_cartesian(xlim = c(0.5 * 10 / MW_GLUCOSE * 0.85,
                           4.0 * 10 / MW_GLUCOSE * 1.03)) +
  labs(title = "SC5314 endpoint OD600 across glucose concentrations",
       x = "[Glucose] (M)",
       y = "OD600 (blank- and dilution-corrected)") +
  theme_bw(base_size = 13) +
  theme(
    panel.border      = element_blank(),
    panel.background  = element_rect(fill = "grey92", colour = NA),
    panel.grid.major  = element_line(colour = "white", linewidth = 0.5),
    panel.grid.minor  = element_blank(),
    axis.line         = element_line(colour = "black")
  )

p_glc
ggsave(out_path("Figure_1_yield_vs_glucose_rich.png"), p_glc, width = 7, height = 5, dpi = 300)
ggsave(out_path("Figure_1_yield_vs_glucose_rich.pdf"), p_glc, width = 7, height = 5)
```

## Session info

```{r session}
sessionInfo()
```
