---
title: "Figure 1: Growth and residual glucose over time in rich and minimal media"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
  cycles: 1000
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(patchwork)
})
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(f) file.path(params$data_dir, f)
out_path <- function(...) file.path(params$out_dir, ...)
```

## Figure legend

**Figure 1 (current figure panels C and D; not described in the legend document, whose panel B
describes the earlier minimal-media growth curves at several glucose concentrations).**
Population growth (OD at 600nm, left axis, dark red) and residual glucose in the medium (mM,
right axis, gold) over time for *C. albicans* SC5314 grown in rich media (C) and minimal media
(D). Mean ± SD.

## Input data

| File | Description |
|---|---|
| `Resubmission_GrowthOD_combined.csv` | Spark plate-reader export, rich-media endpoint OD600 at 0, 6, 24 and 72 h (triplicate columns 1–3, 4–6, 7–9, 10–12); row A = WT SC5314, row G = rich-media blank |
| `20220808_candida_NCs_starve_growth_assay.csv` | Tecan continuous OD600 time course (1000 cycles, ~96 h) of minimal media with varying glucose / nitrogen; one column per well, time in seconds |
| `quiescence_nit+glu_OBI_samplesheet.csv` | Plate layout for the Tecan time course (Well, Sample, Bio_rep, Tech_rep, groups, `C_only`, `N_only`) |
| `Resubmission_GlucoseAssay_Repeat.xlsx` | Glucose assay plate-reader export (absorbance) of culture supernatants at 2, 5, 8, 11, 24, 48, 72 h plus 0/90/180/300 µM standards |
| `glucose_assay_sample_sheet.csv` | Plate layout of the glucose assay (Well, Sample_name, Type, Media, Time, Repeat); wells not listed are unused background wells |

Processing:

* *Rich-media OD*: WT SC5314 (row A); the row-G blank is subtracted per time point; 24 h
  readings are multiplied by 2 and 72 h readings by 4 (dilution before reading); negative values
  clipped to zero.
* *Minimal-media OD*: Tecan time course with outlier wells D2, E2, B6, C6 and D6 removed,
  0.1% glucose condition (`C_only == 2`, samples WT1_8X_C / WT2_8X_C). Mean OD at the cycle
  closest to each of 2, 5, 8, 11, 24, 48 and 72 h; not blank-subtracted.
* *Residual glucose*: plate background = mean of unused wells; standard curve through the origin;
  wells at or below background or below the LOQ (15 × SD of background wells) set to 0 µM;
  well E1 (Minimal 72 h, replicate 2) excluded as a pipetting artefact; dilution factors 4000×
  (rich) and 100× (minimal); 50%-of-median per-group QC gate.

## Minimal-media OD time course

```{r tecan}
samplesheet <- read_csv(data_path("quiescence_nit+glu_OBI_samplesheet.csv"))

# time from seconds to hours
data_od <- read_csv(data_path("20220808_candida_NCs_starve_growth_assay.csv"),
                    skip = 0, n_max = params$cycles)
colnames(data_od)[1] <- "time"
data_od$time <- as.numeric(str_remove(data_od$time, "s")) / 60 / 60
data_od <- data_od[, -98]

# outlier wells
outlier_wells <- c("D2", "E2", "B6", "C6", "D6")
data_od <- data_od %>% select(-all_of(outlier_wells))
```

## Residual glucose

```{r glucose}
fig1b_times <- c(2, 5, 8, 11, 24, 48, 72)

glc_sample_sheet <- read_csv(data_path("glucose_assay_sample_sheet.csv"))

raw_glc <- read_excel(data_path("Resubmission_GlucoseAssay_Repeat.xlsx"),
                      col_names = FALSE, .name_repair = "minimal")
glc_header_row <- which(raw_glc[[1]] == "<>")[1]
glc_plate <- raw_glc[(glc_header_row + 1):(glc_header_row + 8), 1:13]
colnames(glc_plate) <- c("Row", as.character(1:12))

glc_wells <- glc_plate %>%
  pivot_longer(cols = -Row, names_to = "Col", values_to = "Absorbance") %>%
  mutate(Well = paste0(Row, Col), Absorbance = as.numeric(Absorbance)) %>%
  select(Well, Absorbance) %>%
  left_join(glc_sample_sheet, by = "Well")

# plate background = mean of wells not present in the sample sheet
glc_background <- glc_wells %>%
  filter(is.na(Sample_name)) %>%
  summarise(background = mean(Absorbance)) %>%
  pull(background)

glc_wells <- glc_wells %>%
  mutate(Absorbance_corrected = Absorbance - glc_background)

# standard curve constrained through the origin
glc_standards <- glc_wells %>%
  filter(Type == "Standard") %>%
  mutate(Conc_uM = as.numeric(str_extract(Sample_name, "[0-9]+")))

glc_std_fit <- lm(Absorbance_corrected ~ 0 + Conc_uM, data = glc_standards)
glc_slope   <- unname(coef(glc_std_fit)[["Conc_uM"]])

# limit of quantification in corrected-absorbance space (15 x SD of background wells)
glc_bg_wells <- glc_wells %>% filter(is.na(Sample_name))
glc_loq_abs  <- 15 * sd(glc_bg_wells$Absorbance_corrected)

glucose_ts <- glc_wells %>%
  filter(Type == "Experimental") %>%
  mutate(
    Media      = factor(Media, levels = c("Rich", "Minimal")),
    Dilution   = if_else(Media == "Rich", 4000, 100),
    Glucose_uM = (Absorbance_corrected / glc_slope) * Dilution,
    Glucose_uM = if_else(Absorbance_corrected <= 0 |
                           Absorbance_corrected <= glc_loq_abs, 0,
                         pmax(Glucose_uM, 0)),
    Time       = as.numeric(Time)
  ) %>%
  # 50%-of-median QC gate
  group_by(Media, Time) %>%
  mutate(Group_median = median(Glucose_uM),
         Outlier      = Group_median != 0 &
           abs(Glucose_uM - Group_median) / abs(Group_median) > 0.5) %>%
  ungroup()

# well E1 (Minimal 72 h, replicate 2): pipetting artefact
glucose_ts <- glucose_ts %>% filter(Well != "E1")

glucose_ts %>%
  filter(Outlier) %>%
  select(Well, Sample_name, Media, Time, Glucose_uM, Group_median) %>%
  knitr::kable(digits = 0,
               caption = "Glucose-assay wells removed by the 50%-of-median QC gate")

glucose_summary <- glucose_ts %>%
  filter(!Outlier) %>%
  group_by(Media, Time) %>%
  summarise(mean_mM = mean(Glucose_uM) / 1000,
            sd_mM   = sd(Glucose_uM) / 1000,
            n       = n(), .groups = "drop") %>%
  rename(time_h = Time)

knitr::kable(glucose_summary, digits = 3,
             caption = "Residual glucose (mM, dilution-corrected) per time point")
```

## Rich-media OD

```{r rich-od}
rich_lines <- readLines(data_path("Resubmission_GrowthOD_combined.csv"))
rich_hdr   <- grep("^<>,", rich_lines)
rich_raw   <- read.csv(text = rich_lines[rich_hdr:(rich_hdr + 8)],
                       check.names = FALSE, strip.white = TRUE)
rich_plate <- as.matrix(rich_raw[, -1])
rownames(rich_plate) <- trimws(rich_raw[[1]])
storage.mode(rich_plate) <- "double"

# triplicate columns per time point
rich_time_map <- list(`0` = 1:3, `6` = 4:6, `24` = 7:9, `72` = 10:12)

rich_od <- bind_rows(lapply(names(rich_time_map), function(tt) {
  cols     <- rich_time_map[[tt]]
  t        <- as.numeric(tt)
  blank_mu <- mean(rich_plate["G", cols])  # row G = rich-media blank
  dil      <- ifelse(t == 72, 4, ifelse(t == 24, 2, 1))
  corr     <- pmax((rich_plate["A", cols] - blank_mu) * dil, 0)  # row A = WT
  data.frame(time_h = t, od = corr)
})) %>%
  group_by(time_h) %>%
  summarise(mean_od = mean(od), sd_od = sd(od), n = n(), .groups = "drop")

knitr::kable(rich_od, digits = 3,
             caption = "Rich-media WT SC5314 OD600 (blank- and dilution-corrected)")
```

## Minimal-media OD at the assay time points

```{r minimal-od}
minimal_od_full <- data_od %>%
  pivot_longer(cols = -time, names_to = "well", values_to = "od") %>%
  left_join(samplesheet, by = c("well" = "Well")) %>%
  filter(C_only == 2) %>%          # 0.1% glucose minimal media (8X_C)
  group_by(time) %>%
  summarise(mean_od = mean(od), sd_od = sd(od), n = n(), .groups = "drop")

# measurement cycle closest to each target time point
minimal_od <- fig1b_times %>%
  map_dfr(~ minimal_od_full %>%
            slice_min(abs(time - .x), n = 1, with_ties = FALSE) %>%
            transmute(time_h = .x, time_measured = time, mean_od, sd_od, n))

knitr::kable(minimal_od, digits = 3,
             caption = "Minimal-media OD600 at the assay time points (0.1% glucose)")
```

## Plot

Each panel uses a secondary y-axis: glucose is divided by a per-panel scaling factor so it shares
the grid with OD, and the right-hand axis recovers glucose units.

```{r plot, fig.width=11, fig.height=5}
fig1b_od_color  <- "#7B0D1E"  # dark red
fig1b_glc_color <- "#D97706"  # gold

make_fig1b_panel <- function(od_df, glc_df, media_label) {
  od_max  <- max(od_df$mean_od + od_df$sd_od, na.rm = TRUE)
  glc_max <- max(glc_df$mean_mM + glc_df$sd_mM, na.rm = TRUE)
  y_od    <- od_max * 1.12               # top of the OD (left) axis
  f       <- (glc_max * 1.12) / y_od     # glucose (mM) -> OD scale factor

  ggplot() +
    geom_line(data = glc_df, aes(time_h, mean_mM / f),
              color = fig1b_glc_color, linewidth = 1.1) +
    geom_errorbar(data = glc_df,
                  aes(time_h, ymin = (mean_mM - sd_mM) / f,
                      ymax = (mean_mM + sd_mM) / f),
                  color = fig1b_glc_color, width = 2.5, linewidth = 0.7) +
    geom_point(data = glc_df, aes(time_h, mean_mM / f),
               color = fig1b_glc_color, size = 2.6) +
    geom_line(data = od_df, aes(time_h, mean_od),
              color = fig1b_od_color, linewidth = 1.1) +
    geom_errorbar(data = od_df,
                  aes(time_h, ymin = mean_od - sd_od, ymax = mean_od + sd_od),
                  color = fig1b_od_color, width = 2.5, linewidth = 0.7) +
    geom_point(data = od_df, aes(time_h, mean_od),
               color = fig1b_od_color, size = 2.6) +
    scale_x_continuous(breaks = c(0, 12, 24, 36, 48, 60, 72)) +
    scale_y_continuous(
      limits = c(0, y_od),
      sec.axis = sec_axis(~ . * f, name = "Residual glucose (mM)")
    ) +
    labs(title = paste(media_label, "media"), x = "Time (hours)", y = "OD600") +
    theme_gray(base_size = 12) +
    theme(title = element_text(color = "black", size = 12, face = "bold"),
          axis.title.x = element_text(color = "black", size = 10, face = "bold"),
          axis.title.y = element_text(color = fig1b_od_color, size = 10,
                                      face = "bold"),
          axis.title.y.right = element_text(color = fig1b_glc_color, size = 10,
                                            face = "bold"))
}

p_rich    <- make_fig1b_panel(rich_od,
                              filter(glucose_summary, Media == "Rich"), "Rich")
p_minimal <- make_fig1b_panel(minimal_od,
                              filter(glucose_summary, Media == "Minimal"),
                              "Minimal")

fig_growth <- p_rich | p_minimal
fig_growth

ggsave(out_path("Figure_1_growth_and_residual_glucose.png"), fig_growth,
       width = 11, height = 5, dpi = 300)
ggsave(out_path("Figure_1_growth_and_residual_glucose.pdf"), fig_growth,
       width = 11, height = 5)
```

## Session info

```{r session}
sessionInfo()
```
