---
title: "Figure 1: Bud index and DNA content of proliferative and quiescent cells"
author: "Ozan Berk Imir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  data_dir: "data"
  out_dir: "output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(tidyverse)
  library(emmeans)
})
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
data_path <- function(f) file.path(params$data_dir, f)
out_path <- function(...) file.path(params$out_dir, ...)
```

## Figure legend

**Figure 1F.** Bud index quantification (top) and DNA content staining using SYTOX Green and
flow cytometry (bottom). Proliferative = 6 h, quiescent = 72 h; rich and minimal media.

## Input data

| File | Description |
|---|---|
| `PC_Microscopy_analysis.csv` | ImageJ Cell Counter counts per phase-contrast image (Singlet, Doublet, Triplet, Quadruplet+, Ambiguous) for rich and minimal media at 2, 4, 6, 24, 48, 72 h |
| `Ozan_CellCycle_MediaEffect3-analysis.CSV` | Per-sample (per tube/well) gate statistics from the Cytek SYTOX Green cell-cycle run: 2n and 4n % Parent, with 2n / 4n as fractions of (2n + 4n) |
| `Ozan_CellCycle_MediaEffect3-analysis-reformat.CSV` | 2n / 4n fractions per Time × Media × biological replicate (technical replicates at 24–72 h averaged); used for the plot and tests |

Raw FCS files are not distributed. The 2n / 4n gates were set in the Cytek acquisition software
(no gating code); the per-sample gate statistics are in `Ozan_CellCycle_MediaEffect3-analysis.CSV`.

## Bud index (unbudded cells, top)

Unbudded cells = singlets / (singlets + doublets + triplets + quadruplets+), ambiguous cells
excluded; computed per image and summarised as mean ± SE at 6 h and 72 h.

```{r bud-index}
df <- read_csv(data_path("PC_Microscopy_analysis.csv"), show_col_types = FALSE)

quad_col <- grep("^Quadruplet", names(df), value = TRUE)[1]
if (!is.na(quad_col) && quad_col != "Quadruplet") {
  names(df)[names(df) == quad_col] <- "Quadruplet"
}

df <- df %>%
  mutate(across(c(Singlet, Doublet, Triplet, Quadruplet, Time), as.numeric),
         Media = as.factor(Media))

df_stats <- df %>%
  filter(Time %in% c(6, 72)) %>%
  mutate(
    total_nonamb = Singlet + Doublet + Triplet + Quadruplet,
    singlet_pct = ifelse(total_nonamb > 0,
                         Singlet / total_nonamb * 100,
                         NA)
  ) %>%
  group_by(Media, Time) %>%
  summarise(
    mean_pct = mean(singlet_pct, na.rm = TRUE),
    se_pct = sd(singlet_pct, na.rm = TRUE) /
      sqrt(sum(!is.na(singlet_pct))),
    n_images = sum(!is.na(singlet_pct)),
    .groups = "drop"
  ) %>%
  mutate(
    Time = as.factor(Time),
    Media = factor(Media, levels = c("Rich", "Minimal"))
  )

knitr::kable(df_stats, digits = 2, caption = "Unbudded cells (%) per image, mean ± SE")
```

```{r bud-index-plot, fig.width=6, fig.height=4.5}
p_unbudded <- ggplot(df_stats, aes(x = Media, y = mean_pct, fill = Time)) +
  geom_col(position = position_dodge(width = 0.8),
           color = "black", width = 0.65) +
  geom_errorbar(
    aes(ymin = mean_pct - se_pct, ymax = mean_pct + se_pct),
    position = position_dodge(width = 0.8),
    width = 0.25, linewidth = 0.6
  ) +
  scale_fill_manual(
    values = c("6" = "#f29d9a", "72" = "#5fc9d8"),
    labels = c("6" = "6 hours", "72" = "72 hours"),
    name = "Timepoint"
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(0, max(df_stats$mean_pct + df_stats$se_pct) * 1.15),
    labels = function(x) paste0(x, "%")
  ) +
  labs(x = "Media", y = "Unbudded cells (%)",
       title = "Unbudded cells by media and timepoint") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5))

p_unbudded
ggsave(out_path("Figure_1_unbudded_cells.png"), p_unbudded, width = 6, height = 4.5, dpi = 300)
ggsave(out_path("Figure_1_unbudded_cells.pdf"), p_unbudded, width = 6, height = 4.5)
```

No statistical test for the unbudded-cell percentages is performed here.

## DNA content (2n population, bottom)

Check that the reformatted table is the per-biological-replicate mean of the per-sample gate
statistics.

```{r dna-check}
per_sample <- read_csv(data_path("Ozan_CellCycle_MediaEffect3-analysis.CSV"),
                       show_col_types = FALSE)
df_dna <- read_csv(data_path("Ozan_CellCycle_MediaEffect3-analysis-reformat.CSV"),
                   show_col_types = FALSE)

chk <- per_sample %>%
  group_by(Time, Media, `Biological Rep`) %>%
  summarise(`2n_mean` = mean(`2n`), .groups = "drop") %>%
  inner_join(df_dna, by = c("Time", "Media", "Biological Rep"))
stopifnot(nrow(chk) == nrow(df_dna),
          isTRUE(all.equal(chk$`2n_mean`, chk$`2n`, tolerance = 1e-6)))
```

```{r dna}
df_long <- df_dna %>%
  mutate(`2n` = `2n` * 100) %>%
  select(Time, Media, `2n`) %>%
  rename(Proportion = `2n`) %>%
  mutate(Time = as.numeric(Time)) %>%
  filter(Time %in% c(6, 72))

df_summary <- df_long %>%
  group_by(Time, Media) %>%
  summarise(
    mean_2n = mean(Proportion),
    se_2n = sd(Proportion) / sqrt(n()),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    Time = as.factor(Time),
    Media = factor(Media, levels = c("Rich", "Minimal"))
  )

knitr::kable(df_summary, digits = 2, caption = "2n population (%), mean ± SE")
```

```{r dna-plot, fig.width=6, fig.height=4.5}
p_2n <- ggplot(df_summary, aes(x = Media, y = mean_2n, fill = Time)) +
  geom_col(position = position_dodge(width = 0.8),
           color = "black", width = 0.65) +
  geom_errorbar(
    aes(ymin = mean_2n - se_2n, ymax = mean_2n + se_2n),
    position = position_dodge(width = 0.8),
    width = 0.25, linewidth = 0.6
  ) +
  scale_fill_manual(
    values = c("6" = "#f29d9a", "72" = "#5fc9d8"),
    labels = c("6" = "6 hours", "72" = "72 hours"),
    name = "Timepoint"
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(0, max(df_summary$mean_2n + df_summary$se_2n) * 1.15),
    labels = function(x) paste0(x, "%")
  ) +
  labs(x = "Media", y = "2n Population (%)",
       title = "2n Population by media and timepoint") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5))

p_2n
ggsave(out_path("Figure_1_2n_population.png"), p_2n, width = 6, height = 4.5, dpi = 300)
ggsave(out_path("Figure_1_2n_population.pdf"), p_2n, width = 6, height = 4.5)
```

### Statistics

Two-way ANOVA (Media × Time) with Tukey-adjusted post-hoc comparisons, and Welch t-tests with
Bonferroni correction (n = 4 comparisons).

```{r dna-stats}
# Time is numeric here, as in the source analysis
aov_model <- aov(Proportion ~ Media * Time, data = df_long)
summary(aov_model)

emmeans_results <- emmeans(aov_model, pairwise ~ Media * Time, adjust = "tukey")
emmeans_results$contrasts

rich_6  <- df_long %>% filter(Media == "Rich", Time == 6) %>% pull(Proportion)
rich_72 <- df_long %>% filter(Media == "Rich", Time == 72) %>% pull(Proportion)
min_6   <- df_long %>% filter(Media == "Minimal", Time == 6) %>% pull(Proportion)
min_72  <- df_long %>% filter(Media == "Minimal", Time == 72) %>% pull(Proportion)

comparisons <- list(
  "Rich 6h vs Rich 72h"       = t.test(rich_6, rich_72),
  "Minimal 6h vs Minimal 72h" = t.test(min_6, min_72),
  "Rich 6h vs Minimal 6h"     = t.test(rich_6, min_6),
  "Rich 72h vs Minimal 72h"   = t.test(rich_72, min_72)
)

tibble(comparison = names(comparisons),
       p = map_dbl(comparisons, "p.value"),
       p_adj_bonferroni = p.adjust(p, method = "bonferroni", n = 4)) %>%
  knitr::kable(digits = 4)
```

## Session info

```{r session}
sessionInfo()
```
